%% paper.tex — Testing for Simplicial Geometry
%% arXiv preprint targeting Random Structures and Algorithms (Wiley)
%% Compile: cd simplicial-latent-geometry/my_theorems && pdflatex paper && bibtex paper && pdflatex paper && pdflatex paper

\documentclass[reqno,12pt]{amsart}

%% ── Encoding & fonts ─────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}

%% ── Mathematics ──────────────────────────────────────────────────────────────
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}          % \coloneqq, dcases

%% ── Bibliography ─────────────────────────────────────────────────────────────
\usepackage[numbers,sort&compress]{natbib}

%% ── Hyperlinks ───────────────────────────────────────────────────────────────
\usepackage[dvipsnames,svgnames,x11names]{xcolor}
\usepackage[colorlinks=true,
            linkcolor=RoyalBlue4,
            citecolor=RoyalBlue4,
            urlcolor=NavyBlue]{hyperref}
\usepackage{url}

%% ── Page geometry ────────────────────────────────────────────────────────────
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}

%% ── Enumitem for tighter lists ───────────────────────────────────────────────
\usepackage{enumitem}

%% ─────────────────────────────────────────────────────────────────────────────
%%  MACROS
%% ─────────────────────────────────────────────────────────────────────────────

%% Models and statistics
\newcommand{\tauf}{\tau_{\mathrm{f}}}
\newcommand{\PC}{\mathrm{2PC}}
\newcommand{\Cech}{\ifmmode\text{\v{C}ech}\else\v{C}ech\fi}  % Čech (text+math)
\newcommand{\RGG}{\mathrm{RGG}}
\newcommand{\TV}{\mathrm{TV}}

%% Named quantities
\newcommand{\geomCov}{\mathrm{geomCov}}
\newcommand{\fillingP}{\mathrm{fillingProb}}
\newcommand{\matchR}{\mathrm{matchRadius}}

%% Sets and spaces
\newcommand{\TT}{\mathbb{T}}         % flat torus
\newcommand{\AddCircle}{\mathbb{T}^1}% AddCircle(1) = ℝ/ℤ, the unit circle
\newcommand{\RR}{\mathbb{R}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\Sph}{\mathbb{S}}

%% Probability
\newcommand{\EE}{\mathbb{E}}
\newcommand{\PP}{\mathbb{P}}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}

%% Lean annotation (gray box for theorem names)
\newcommand{\lean}[1]{\textup{\texttt{\small #1}}}
\newcommand{\leanverified}[1]{\textup{(Lean: \texttt{#1}, 0 sorries)}}


%% ─────────────────────────────────────────────────────────────────────────────
%%  THEOREM ENVIRONMENTS
%% ─────────────────────────────────────────────────────────────────────────────

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{observation}[theorem]{Observation}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{conjecture}[theorem]{Conjecture}

%% ─────────────────────────────────────────────────────────────────────────────
%%  DOCUMENT
%% ─────────────────────────────────────────────────────────────────────────────

\begin{document}

\title[Testing for Simplicial Geometry]{Testing for Simplicial Geometry: A Signed Triangle
  Statistic for the \v{C}ech--2PC Phase Transition}

\author[Goh and Cook]{David Goh}
\address{Department of Computer Science, University of Toronto}
\email{daveed@cs.toronto.edu}

\author{Nicholas A.\ Cook}
\address{Department of Mathematics, Duke University}

\date{}

\subjclass[2020]{%
  Primary 05C80 (random graphs, generic models);
  Secondary 55U10 (simplicial sets and complexes),
  62H30 (classification and discrimination; cluster analysis).}

\keywords{random simplicial complex, geometry detection, Čech complex, signed triangle statistic,
  phase transition, Lean 4, formal verification}

\begin{abstract}
Random simplicial complexes are natural higher-dimensional analogues of random graphs, arising
in topological data analysis and the study of networks with higher-order structure. A basic
question in statistical inference on such objects is whether an observed complex reflects
latent geometric organization---in which the vertices have positions in some ambient space
that drive the connection structure---or whether its triangle structure appeared by chance,
with no underlying geometry. We study a precise formulation of this geometry detection
problem: given an observed 2-dimensional simplicial complex on $n$ vertices, determine whether
it arose from the \emph{2-parameter complex} $\PC(n,p,q)$---in which edges and triangle fills
are independent---or from the \v{C}ech complex of $n$ points sampled uniformly on the flat
torus $\TT^d$, with radius chosen to match the edge probability~$p$.

We introduce the \emph{doubly-signed filled triangle statistic} $\tauf$, centered with respect
to both edge and fill indicators. Under $\PC$, $\tauf$ has mean zero and variance $\Theta(n^3)$.
Under the \v{C}ech model, $\tauf$ has mean $\binom{n}{3}\cdot\geomCov(p,d)$, where
$\geomCov(p,d)\to 0$ as $d\to\infty$. There is an explicit sharp threshold
$d^*(p) = |\log p|/\log(3/2)$: for $d \geq d^*(p)$, every triple of points on $\TT^d$ fills
and $\geomCov(p,d) = 0$ \emph{exactly}, so $\tauf$-based detection is impossible; for $d < d^*(p)$,
$\geomCov(p,d) > 0$ and detection succeeds when $n^{3/2}\cdot\geomCov(p,d)\to\infty$.
All moment results, detection lower bound, the phase transition $\geomCov(p,d)=0$ for $d\geq d^*(p)$,
and the limit $\geomCov(p,d)\to 0$ are formally verified in Lean~4 with
Mathlib~\cite{Mathlib2020}, with no remaining sorries in the active proof chain.
\end{abstract}

\maketitle

%% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{Background and Motivation}

Topological data analysis (TDA) infers the shape of data by building simplicial complexes on
point clouds and measuring their homology. The foundational pipeline---sample $n$ points, form
the \v{C}ech or Vietoris-Rips complex at radius $r$, read off Betti numbers or persistent
homology---rests on the assumption that observed topological features reflect genuine geometry
in the underlying data. But any finite point cloud will produce some triangles and higher faces
by chance; the practical question is whether those topological features are signal or noise.

Kahle~\cite{Kahle2011} systematically studied the expected topology of \v{C}ech and
Vietoris-Rips complexes built on i.i.d.\ random points in $\RR^d$, establishing connectivity
and component-count thresholds as higher-dimensional analogues of results for random geometric
graphs. A central motivation, surveyed in Bobrowski and Kahle~\cite{Bobrowski2018}, is the
need for probabilistic null hypotheses against which TDA test statistics can be calibrated;
the random geometric complex on i.i.d.\ uniform points is the natural geometric alternative
to such a null.

The simplest null model for 2-dimensional complexes is the \emph{2-parameter complex}
$\PC(n,p,q)$: place $n$ vertices, insert each edge independently with probability $p$, and
fill each completed triangle independently with probability $q$ (see also the multi-parameter
model of Kahle~\cite{Kahle2016}). This encodes no geometry---all events are mutually
independent. The geometric alternative is the \emph{\v{C}ech complex} $\Cech(n,r,d)$: sample
$n$ points uniformly on the flat torus $\TT^d$, connect two points whenever they lie within
distance $r$, and fill a triangle whenever the three associated sup-norm balls of radius $r$
share a common point (the nerve criterion). The fundamental detection question is: given only
the observed simplicial complex, can one tell which model generated it?

\subsection{Related Work}\label{sec:related}

\paragraph{Geometry testing for random graphs.}
Bubeck, Ding, Eldan, and R\'acz~\cite{Bubeck2016} introduced the signed triangle statistic
for testing $G(n,p)$ vs.\ $\RGG(n,p,d)$ on $\Sph^{d-1}$, establishing the $d\asymp n^3$
phase transition: random geometric graphs are detectable when $d\ll n^3$ and
indistinguishable from Erd\H{o}s--R\'enyi when $d\gg n^3$. Bangachev and
Bresler~\cite{BangachevBresler2024} studied the low-degree Fourier coefficients of random
geometric graphs on $\Sph^{d-1}$, showing that the signed triangle statistic is
computationally optimal for this testing problem and characterizing the Fourier coefficient
behavior across the full range of densities. Liu, Mohanty, Schramm, and Yang~\cite{Liu2022}
obtained sharp detection thresholds for sparse random geometric graphs via the low-degree
polynomial framework; their methods provide a model for sharpening the SNR condition below the
detection threshold. Brennan, Bresler, and Huang~\cite{Brennan2022} extended BDER to
anisotropic latent geometries; a complementary line due to Litvak, Michielan, and
Stegehuis~\cite{Litvak2022} shows that triangle-count statistics are insufficient to detect
\emph{hyperbolic} geometry in networks, underscoring that the power of triangle-based tests
is geometry-dependent. Perkins~\cite{Perkins2024} surveys recent progress on sharp
thresholds in random structures.
Our work carries this framework into the simplicial setting, where the presence of fill
indicators introduces an additional layer of geometric dependence beyond edge structure alone.
The key qualitative difference from BDER is that their detection threshold
$d^*\asymp n^3$ grows with~$n$: edge correlations on $\Sph^{d-1}$ remain positive for every
finite~$d$, so the threshold is controlled entirely by the SNR condition
$n^3\cdot\mathrm{cov}^2\to\infty$. Our simplicial analogue, by contrast, has a
\emph{finite, $n$-independent} critical dimension: the fill layer introduces a complete
collapse of geometric signal at a computable threshold, making detection impossible
above it regardless of sample size (Theorem~\ref{thm:phase-transition}).

\paragraph{Hypergraph detection.}
Dhawan, Mao, and Wein~\cite{DhawanMaoWein2023} study detection of planted dense
subhypergraphs via low-degree polynomials in $r$-uniform hypergraphs, establishing sharp
thresholds in the log-density regime. Their setting is combinatorial (planted structure in an
Erd\H{o}s--R\'enyi null) rather than geometric, but the low-degree framework and the
connection between hyperedge counting statistics and the structure of test polynomials is
closely related to our context. We view their analysis as the combinatorial cousin of our
geometric detection problem: the 2PC null here plays the role of their Erd\H{o}s--R\'enyi
hypergraph, and the \v{C}ech complex plays the role of their planted dense substructure,
but the source of dependence (latent torus positions rather than a planted clique) is
qualitatively different and yields a different class of tests.

\paragraph{Random geometric complexes and TDA.}
Kahle~\cite{Kahle2011} established homology thresholds for \v{C}ech and Vietoris-Rips
complexes on random points in $\RR^d$, motivated by the need for null models in TDA.
Kahle~\cite{Kahle2016} surveys the multi-parameter model $X(n;p_1,p_2,\ldots)$ (of which
our $\PC(n,p,q)$ is a special case) and the connection to persistent homology.
Bobrowski and Kahle~\cite{Bobrowski2018} survey random geometric complexes, explicitly
connecting the topology of such complexes to null hypotheses for topological statistics.
Tem\v{c}inas, Nanda, and Reinert~\cite{Temcinas2025} develop goodness-of-fit tests for
the multi-parameter random simplicial complex $\mathbf{X}(n,\mathbf{p})$ in the dense
regime, using multivariate CLTs for subcomplex counts proved via Stein's method.
Their setting and ours are complementary: they test a single null
$\mathbf{X}(n,\mathbf{p})$ against unspecified alternatives (one-sample goodness-of-fit),
while we formulate a two-sample detection problem between a combinatorial null
($\PC$) and a geometric alternative ($\Cech$). A technical point of contact is their
observation that raw subcomplex counts in $\mathbf{X}(n,\mathbf{p})$ have rank-one
limiting covariance, which motivates their switch to critical simplex counts from
discrete Morse theory; our doubly-signed $\tauf$ sidesteps this degeneracy directly
by centering both edge and fill indicators, yielding non-degenerate
$\Theta(n^3)$ variance under $\PC$ (Proposition~\ref{prop:moments-twopc}).
Tem\v{c}inas et al.\ also empirically demonstrate that their test can distinguish
$\mathbf{X}(n,\mathbf{p})$ from selected geometric simplicial complex models;
the present work gives a theoretical analogue with sharp phase transitions on
$\TT^d$.

\subsection{Main Results and Contributions}\label{sec:mainresults}

The key tool is the \emph{doubly-signed filled triangle statistic}
\begin{equation}\label{eq:tauf}
  \tauf = \sum_{\{i,j,k\}\in\binom{[n]}{3}}
    \Bigl[\prod_{e\in\partial\{i,j,k\}}(A_e - p)\Bigr](F_{ijk} - q),
\end{equation}
centered with respect to both edge and fill variability (Definition~\ref{def:tauf}).
The quantity $\geomCov(p,d)$ (Definition~\ref{def:geomcov}) is the per-triangle \v{C}ech-model
expectation of this doubly-centered product; it is positive when fills carry geometric
information and collapses to zero when geometry becomes negligible.

The paper has two main results. The first (Theorem~\ref{thm:detection}) is a detection
bound: for any fixed dimension $d$ at which $\geomCov(p,d) > 0$, the test that thresholds
$\tauf$ achieves asymptotic power $1$ as $n\to\infty$, so $\TV(\PC,\Cech)\to 1$. The
signal-to-noise calculation (Lemma~\ref{lem:snr}) shows that the underlying detection
criterion is $n^{3/2}\cdot\geomCov(p,d)\to\infty$; under $d$ fixed this is automatic from
positivity of $\geomCov$.

The second (Theorem~\ref{thm:phase-transition}) is a sharp phase transition at a finite,
$n$-independent critical dimension
\[
  d^*(p) \;=\; \frac{|\log p|}{\log(3/2)}\;\approx\;2.47\,|\log p|.
\]
For $d < d^*(p)$, $\geomCov(p,d) > 0$ and detection succeeds; for $d \geq d^*(p)$,
$\geomCov(p,d) = 0$ exactly (not merely in the limit) and no $\tauf$-based test can
distinguish the models. The collapse is driven by a 1D Helly argument: once
$r(p,d) = p^{1/d}/2 \geq 1/3$, every three sup-norm balls on the torus share a common
point, so every triple fills deterministically and the fill indicator carries no signal.

\medskip
\noindent\textbf{Contributions.}
\begin{enumerate}[label=(\arabic*)]
\item \textbf{The doubly-signed statistic} (Definition~\ref{def:tauf},
  equation~\eqref{eq:tauf}).
  The natural simplicial analogue of the BDER signed triangle statistic, centering both edge
  and fill variability.

\item \textbf{Moment computations} (Propositions~\ref{prop:moments-twopc}
  and~\ref{prop:moments-cech}).
  Under $\PC(n,p,q)$, $\EE[\tauf] = 0$ and
  $\Var[\tauf] = \binom{n}{3}p^3(1-p)^3 q(1-q)$.
  Under $\Cech(n,r,d)$ with matched parameters,
  $\EE[\tauf] = \binom{n}{3}\cdot\geomCov(p,d)$.

\item \textbf{Phase transition and finite critical dimension}
  (Definition~\ref{def:critical-dim}, Theorem~\ref{thm:phase-transition}).
  For $d \geq d^*(p)$, the matched ball radius exceeds $1/3$, every triple fills by a
  1D Helly argument (Lemma~\ref{lem:helly}), and $\geomCov(p,d) = 0$ exactly.

\item \textbf{Lean~4 verification.}
  The moment results, detection lower bound, phase transition (both regimes), and
  $\geomCov(p,d) \to 0$ are formally verified in Lean~4 with Mathlib, with no remaining
  sorries in the active proof chain; see Appendix~\ref{app:lean} for the full catalog.
\end{enumerate}

\subsection{Paper Organization}

Section~\ref{sec:models} defines the models and parameter matching, and establishes the key
asymptotic properties of the matched radius and geometric covariance
(Lemmas~\ref{lem:radius-asymptotics} and~\ref{lem:helly}).
Section~\ref{sec:statistic} introduces $\tauf$ and proves its moments under both models.
Section~\ref{sec:detection} carries out the detection analysis and proves
Theorems~\ref{thm:detection} and~\ref{thm:phase-transition}; the logical flow is
Propositions~\ref{prop:moments-twopc} and~\ref{prop:moments-cech} feeding into
Lemma~\ref{lem:snr}, which implies Theorem~\ref{thm:detection}; part~(a) of
Theorem~\ref{thm:phase-transition} then follows from Theorem~\ref{thm:detection} and
Lemma~\ref{lem:snr}, while part~(b) follows from
Lemmas~\ref{lem:radius-asymptotics} and~\ref{lem:helly}.
Section~\ref{sec:discussion} discusses implications and open problems.
Appendix~\ref{app:lean} is a catalog of all formally verified Lean~4 results;
in-text annotations of the form \lean{lemma\_name} throughout the paper link to the
corresponding entries there.

\subsection{AI-Assisted Mathematical Development}\label{sec:ai}

The development of this paper was assisted by two AI tools.
\textbf{Claude Code}~\cite{ClaudeCode2025} (Anthropic) contributed at both the mathematical
and formal levels: it assisted with the formulation of the doubly-signed statistic~$\tauf$,
the BDER analogy as a guiding framework, and the phase transition analysis, as well as proof
architecture, tactic selection, and Lean~4 syntax over approximately~45 formalization rounds.
The \textbf{Aristotle API}~\cite{Aristotle2025} (Harmonic.fun) performed automated formal
proof completion for individual lemmas; each returned proof was reviewed and cherry-picked
into the codebase by the authors.

This work follows the approach of Jang and Ryu~\cite{JangRyu2026}, who employed ChatGPT in
the \emph{discovery} of their proof of point convergence of Nesterov's accelerated gradient
method. As in that work, AI assistance extended beyond the formal verification layer into
mathematical exploration. The authors are responsible for all mathematical content.

%% ─────────────────────────────────────────────────────────────────────────────
\section{Models and Parameter Matching}\label{sec:models}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{Two Random Models of Simplicial Complexes}

\begin{definition}[2-parameter complex]\label{def:twopc}
The \emph{2-parameter complex} $\PC(n,p,q)$ is the random 2-dimensional simplicial complex
on vertex set $[n] = \{1,\ldots,n\}$ constructed as follows: each edge $\{i,j\}$ is included
independently with probability $p\in(0,1)$; each 3-clique $\{i,j,k\}$ (three vertices with
all three edges present) is filled to a 2-simplex independently with probability $q\in(0,1)$.
\end{definition}

The 2PC is the natural null model for random simplicial complexes: it places filled triangles
at each available 3-cycle with no imposed geometry. Under $\PC$, edge indicators $A_{ij}$
and fill indicators $F_{ijk}$ are all mutually independent.

\begin{definition}[Flat torus]\label{def:torus}
Let $d\geq 1$. The \emph{flat torus} $\TT^d$ is the product space $(\RR/\ZZ)^d$ equipped
with the sup-norm metric
\[
  d_\infty(x,y) = \max_{i\leq d}\lvert x_i - y_i\rvert_\ZZ,
  \quad \lvert t\rvert_\ZZ = \min_{k\in\ZZ}\lvert t - k\rvert.
\]
Under this metric, the ball of radius $r$ centered at any point has volume $(2r)^d$,
provided $r \leq 1/2$.
\end{definition}

\begin{definition}[\v{C}ech complex]\label{def:cech}
The \emph{\v{C}ech complex} $\Cech(n,r,d)$ is constructed by sampling $n$ points
$x_1,\ldots,x_n$ uniformly and independently on $\TT^d$. The edge $\{i,j\}$ is included
when $d_\infty(x_i,x_j)\leq r$. The fill indicator $F_{ijk}$ records whether the three
sup-norm balls $B(x_i,r)$, $B(x_j,r)$, $B(x_k,r)$ share a common point (the nerve
criterion):
\[
  F_{ijk} = \mathbf{1}\!\left[\bigcap_{\ell\in\{i,j,k\}} B(x_\ell,r) \neq \emptyset\right].
\]
\end{definition}

\begin{remark}[Two conventions for the \v{C}ech fill]\label{rem:fill-conventions}
There are two natural ways to define the fill indicator on the flat torus with the
sup-norm. The \emph{nerve-only} convention (used here, and in the Lean formalization)
takes $F_{ijk}$ to depend only on triple ball intersection. The \emph{downward-closed}
convention takes $F_{ijk} = \mathbf{1}[A_{ij}A_{ik}A_{jk}=1]\cdot
\mathbf{1}[\cap_\ell B(x_\ell,r)\neq\emptyset]$, which enforces the simplicial-complex
closure property (every filled 2-face has its boundary edges). Since our edges are
defined by the stricter condition $d_\infty\leq r$ (rather than ball-pair intersection
$d_\infty\leq 2r$), the two conventions disagree: a triple intersection only implies
pairwise distance $\leq 2r$. We adopt the nerve-only convention because it yields a
cleaner joint law $(A_{ij}, A_{ik}, A_{jk}, F_{ijk})$ whose moment calculations factor
cleanly. The signed statistic $\tauf$ and the detection analysis go through under either
convention with only superficial changes.
\end{remark}

\subsection{Parameter Matching}

To compare $\PC$ and $\Cech$ on equal footing, we match the marginal edge probability $p$
and the marginal fill probability $q$.

\begin{definition}[matchRadius]\label{def:matchradius}
Given $p\in(0,1)$ and $d\geq 1$, the \emph{matched radius} is
\begin{equation}\label{eq:matchradius}
  r(p,d) = \frac{p^{1/d}}{2}.
\end{equation}
This is the unique $r\in(0,1/2]$ satisfying $(2r)^d = p$, i.e., the sup-norm ball of radius
$r$ has volume exactly $p$.
\end{definition}

With this choice, points $x_i, x_j$ sampled uniformly on $\TT^d$ satisfy
$\PP[d_\infty(x_i,x_j)\leq r] = p$. The formula~\eqref{eq:matchradius} is formally verified
in Lean as \lean{matchRadius\_spec} \leanverified{matchRadius\_spec}.

\begin{definition}[fillingProb]\label{def:fillingprob}
Let $r = r(p,d)$. The \emph{filling probability} is the marginal probability that the
nerve condition holds on three iid uniform points:
\[
  \fillingP(p,d)
  = \PP\!\left[\bigcap_{\ell=1}^{3} B(x_\ell, r)\neq\emptyset\right],
  \qquad x_1,x_2,x_3\sim_{\mathrm{iid}}\mathrm{Unif}(\TT^d).
\]
\end{definition}

We set $q = \fillingP(p,d)$ so that both models share the same marginal fill probability.
Under this matching, any distributional difference in $\tauf$ reflects geometric dependence,
not marginal mismatch.

We adopt the following vocabulary throughout. A triple $\{i,j,k\}$ is an \emph{empty
triangle} if all three edges $A_{ij},A_{ik},A_{jk}$ are present but $F_{ijk} = 0$ (in the
\v{C}ech model, this means the three pairwise sup-norm balls meet pairwise but have no
common point); it is a \emph{filled triangle} if additionally $F_{ijk} = 1$. Under 2PC,
empty and filled triangles are both random events (with conditional probabilities $1-q$
and $q$ given the three edges). The phase transition in Section~\ref{sec:detection}
is driven by the disappearance of empty triangles under $\Cech$: above the critical
dimension, every triple of present edges automatically fills, eliminating the fill
indicator's random variability.

\begin{definition}[geometricCov]\label{def:geomcov}
The \emph{geometric covariance} $\geomCov(p,d)$ is the per-triangle contribution to
$\EE[\tauf]$ under $\Cech(n,r,d)$ with matched parameters:
\begin{equation}\label{eq:geomcov-def}
  \geomCov(p,d)
  = \EE_{\Cech}\!\bigl[(A_{ij}-p)(A_{ik}-p)(A_{jk}-p)(F_{ijk}-q)\bigr],
\end{equation}
where the expectation is over a single triangle $\{i,j,k\}$ drawn uniformly under
$\Cech(n,r,d)$ with $r = r(p,d)$ and $q = \fillingP(p,d)$.
\end{definition}

The quantity $\geomCov(p,d)$ has two distinguishable pieces. The three factors
$(A_e - p)$ form the \emph{BDER signed edge product}: this is the multiplicand that
BDER~\cite{Bubeck2016} use for graph detection on $\Sph^{d-1}$, and it isolates the
pairwise-edge correlations induced by shared latent coordinates. On its own, it is
sensitive only to graph-level geometry (how close two edges become when they share a
vertex). The additional factor $(F_{ijk} - q)$ is specific to the simplicial setting:
it centers the nerve-criterion indicator, encoding triple-overlap information---whether
the three sup-norm balls share a common point, not merely intersect pairwise. Multiplying
and centering all four yields a per-triangle quantity that vanishes whenever any one of
the indicators decouples from the rest. Above the critical dimension, $F$ becomes
deterministic given the edges and the fill layer ceases to carry information, collapsing
$\geomCov$ to zero (Theorem~\ref{thm:phase-transition}(b))---a collapse that is
\emph{invisible} at the level of the BDER signed edge product alone, since that
quantity remains positive for every finite~$d$ on every torus.

\subsection{Asymptotic Properties of the Matched Parameters}

The phase transition established in Section~\ref{sec:detection} is driven by a purely
geometric fact about the sup-norm torus, which we isolate first. Once this 1D Helly-type
lemma is in hand, the asymptotic collapse of $\geomCov(p,d)$ follows quickly by matching
radii.

\begin{lemma}[Three-arc intersection]\label{lem:helly}
Let $r \geq 1/3$ and let $x_1,x_2,x_3\in\TT^d$ be arbitrary. Then
$B(x_1,r)\cap B(x_2,r)\cap B(x_3,r)\neq\emptyset$.
\hfill \leanverified{addCircle\_three\_balls\_intersect'}
\end{lemma}

\begin{proof}
By the product structure $\TT^d = (\AddCircle)^d$, it suffices to prove the statement in
each coordinate separately: three sup-norm balls in $\TT^d$ intersect iff their projections
onto each $\AddCircle$-coordinate intersect. So it suffices to treat the case $d = 1$.

On $\AddCircle = \RR/\ZZ$ with the sup-norm, the ball $B(x_i,r)$ is a closed arc of length
$2r$, and its complement $B(x_i,r)^c$ is an open arc of length $1-2r$.

\textit{Case $r > 1/3$:} The three open complements have lengths $1-2r < 1/3$ each, with
total length $3(1-2r) < 1$. Since they cannot cover $\AddCircle$ (circumference~1), there
exists $y \notin \bigcup_i B(x_i,r)^c$, i.e., $y \in \bigcap_i B(x_i,r)$.

\textit{Case $r = 1/3$:} The three complements have length exactly $1/3$ each. If they do
not cover $\AddCircle$, a common point of the three balls exists as above. If they do cover
$\AddCircle$, they must be pairwise disjoint open arcs forming a partition (up to finitely
many boundary points). Denote the three boundary points (endpoints of the complementary arcs)
by $c_1, c_2, c_3$. Each $c_j$ is an endpoint of exactly two complementary arcs, hence lies
in the closure of two of the balls and in the interior of the third. In particular,
$c_1, c_2, c_3 \in B(x_1, 1/3) \cap B(x_2, 1/3) \cap B(x_3, 1/3)$, so the triple
intersection is nonempty.

In both cases, $\bigcap_i B(x_i, r) \neq \emptyset$.
\end{proof}

With Lemma~\ref{lem:helly} in hand, we turn to the matched parameters.

\begin{lemma}[Radius asymptotics]\label{lem:radius-asymptotics}
Fix $p\in(0,1)$.
\begin{enumerate}[label=(\alph*)]
\item $r(p,d) = p^{1/d}/2\to 1/2$ as $d\to\infty$; the matched radius increases
  monotonically toward the torus diameter.
  \hfill \leanverified{matchRadius\_tendsto\_half}
\item $\geomCov(p,d)\to 0$ as $d\to\infty$.
  \hfill \leanverified{geometricCov\_tendsto\_zero}
\end{enumerate}
\end{lemma}

\begin{proof}
\textit{Part~(a).} Since $p\in(0,1)$, we have $\log p < 0$, so
$\log r(p,d) = (\log p)/d - \log 2 \to -\log 2$ as $d\to\infty$, giving $r(p,d)\to 1/2$.
Monotonicity holds because $p^{1/d}$ is increasing in $d$ for $p\in(0,1)$.

\textit{Part~(b).} The critical dimension $d^*(p) = |\log p|/\log(3/2)$ is finite.
We claim that for all integer $d \geq \lceil d^*(p)\rceil$, $\geomCov(p,d) = 0$ exactly;
the limit follows immediately since the sequence is eventually zero.

To see this, fix $d \geq d^*(p)$. Then $d \log(3/2) \geq |\log p| = \log(1/p)$, so
$(3/2)^d \geq 1/p$, giving $p^{1/d} \geq 2/3$ and $r(p,d) = p^{1/d}/2 \geq 1/3$.
By Lemma~\ref{lem:helly}, any three balls of radius $r \geq 1/3$ on $\TT^d$ have
nonempty triple intersection, so every triple of edge-present points fills and
$\fillingP(p,d) = 1$, giving $q = 1$. Consequently, on the event that all three edges
are present, $F_{ijk} = 1 = q$; on the event that some edge is absent, $F_{ijk} = 0$ but
the accompanying signed edge factors multiply out to zero in expectation under
translation invariance. A careful algebraic decomposition (formalized in Lean as
\lean{geometricCov\_eventually\_zero}) yields $\geomCov(p,d) = 0$ exactly. See
Theorem~\ref{thm:phase-transition}(b) for the detection-side consequence. \qedhere
\end{proof}

%% ─────────────────────────────────────────────────────────────────────────────
\section{The Doubly-Signed Statistic}\label{sec:statistic}
%% ─────────────────────────────────────────────────────────────────────────────

Two natural simplicial analogues of the BDER signed triangle statistic suggest themselves,
and neither is quite right. Centering only the edge indicators (as BDER does for graphs)
gives a statistic sensitive to graph geometry but blind to fill structure; centering only
the fill indicator leaves the edge layer unsigned, so the expectation under $\PC$ picks up
a systematic bias from the correlation between edges and fills. The construction below
centers all four indicators simultaneously: we replace each $A_e$ by $A_e - p$ and
$F_{ijk}$ by $F_{ijk} - q$. This doubly-centered product vanishes in expectation under
$\PC$ (by independence) and isolates the geometric dependence introduced by the
\v{C}ech model.

Let $A_{ij}\in\{0,1\}$ denote the indicator that edge $\{i,j\}$ is present, and let
$F_{ijk}\in\{0,1\}$ denote the indicator that triangle $\{i,j,k\}$ is filled (zero if any
edge is absent). Fix parameters $p,q$ from the null model.

\begin{definition}[Doubly-signed statistic]\label{def:tauf}
The \emph{doubly-signed filled triangle statistic} is
\begin{equation}\label{eq:tauf-def}
  \tauf = \sum_{\{i,j,k\}\in\binom{[n]}{3}}
    \Bigl[\prod_{e\in\partial\{i,j,k\}}(A_e - p)\Bigr](F_{ijk} - q),
\end{equation}
where $\partial\{i,j,k\} = \bigl\{\{i,j\},\{i,k\},\{j,k\}\bigr\}$ denotes the three
boundary edges.
\end{definition}

The factor $\prod_{e\in\partial t}(A_e - p)$ is the BDER signed edge product for
triangle~$t$; the factor $(F_t - q)$ centers the fill indicator. The product is doubly
centered, vanishing in expectation whenever any of the four indicators is independent of the
rest. We write $Z_t = (A_{ij}-p)(A_{ik}-p)(A_{jk}-p)(F_{ijk}-q)$ for the per-triangle
summand, so $\tauf = \sum_t Z_t$.

We first compute the moments under the null model, where full independence reduces the
calculation to products of centered indicators.

\begin{proposition}[Moments under 2PC]\label{prop:moments-twopc}
Under $\PC(n,p,q)$:
\begin{enumerate}[label=(\alph*)]
\item $\EE[\tauf] = 0$.
\item $\Var[\tauf] = \binom{n}{3}\cdot p^3(1-p)^3\cdot q(1-q)$.
  In particular, $\Var[\tauf\mid\PC] = \Theta(n^3)$.
\end{enumerate}
\hfill \leanverified{moments\_twoParam\_signed}
\end{proposition}

\begin{proof}
\textit{Part~(a).} Under $\PC$, all four indicators $A_{ij}, A_{ik}, A_{jk}, F_{ijk}$ are
mutually independent. Hence
\[
  \EE[Z_t] = \EE[A_{ij}-p]\cdot\EE[A_{ik}-p]\cdot\EE[A_{jk}-p]\cdot\EE[F_{ijk}-q] = 0,
\]
since $\EE[A_{ij}-p] = 0$. Summing over all $\binom{n}{3}$ triangles gives $\EE[\tauf]=0$.

\textit{Part~(b), off-diagonal terms.} For distinct triangles $t\neq t'$, we show
$\EE[Z_t Z_{t'}] = 0$. Under $\PC$, all edge and fill indicators are mutually independent.
Any edge $e \in \partial t \setminus \partial t'$ that appears in $t$ but not $t'$ contributes
an independent factor $(A_e - p)$ to $Z_t$ which is independent of $Z_{t'}$; since
$\EE[A_e-p]=0$, the product $\EE[Z_t Z_{t'}]=0$. The same argument applies if $t,t'$
share all three edges (impossible for distinct triangles) or if the fill $F_t$ is unshared.
More precisely: whenever any indicator in the product $Z_t Z_{t'}$ appears in exactly one of
$Z_t, Z_{t'}$, it contributes a mean-zero factor independent of the rest, killing the
expectation. One checks case by case that for any two distinct triangles (sharing 0, 1, 2, or
3 vertices), at least one such unshared indicator exists.

\textit{Part~(b), diagonal terms.} By mutual independence in $\PC$,
\[
  \Var[Z_t] = \EE[Z_t^2]
  = \EE[(A_{ij}-p)^2]\cdot\EE[(A_{ik}-p)^2]\cdot\EE[(A_{jk}-p)^2]\cdot\EE[(F_{ijk}-q)^2]
  = p(1-p)\cdot p(1-p)\cdot p(1-p)\cdot q(1-q).
\]
Since the $\binom{n}{3}$ diagonal terms are uncorrelated, summing gives
$\Var[\tauf] = \binom{n}{3}\cdot p^3(1-p)^3 q(1-q)$.
\end{proof}

Under the \v{C}ech model the three vertices of a triangle are coupled through their
shared positions on $\TT^d$, and covariance across triangles is nontrivial only when
two triangles share an edge.

\begin{proposition}[Moments under \v{C}ech]\label{prop:moments-cech}
Under $\Cech(n,r,d)$ with $r = r(p,d)$ and matched $q = \fillingP(p,d)$:
\begin{enumerate}[label=(\alph*)]
\item $\EE[\tauf] = \binom{n}{3}\cdot\geomCov(p,d)$.
  \hfill \leanverified{moments\_cech\_signed}
\item $\Var[\tauf] = O(n^4)$.
\end{enumerate}
\end{proposition}

\begin{proof}
\textit{Part~(a).}
Write $Z_t = (A_{ij}-p)(A_{ik}-p)(A_{jk}-p)(F_{ijk}-q)$ for triangle $t = \{i,j,k\}$.
Under the \v{C}ech model, the joint law of $(A_{ij}, A_{ik}, A_{jk}, F_t)$ is determined by
the triple $(x_i,x_j,x_k)$ drawn uniformly on $(\TT^d)^3$. By definition of
$\geomCov(p,d)$~\eqref{eq:geomcov-def}, we have $\EE[Z_t] = \geomCov(p,d)$ for each
triangle. Since all $\binom{n}{3}$ triangles are identically distributed,
$\EE[\tauf] = \binom{n}{3}\cdot\geomCov(p,d)$.

\textit{Part~(b).}
Expand $\Var[\tauf] = \sum_{t,t'} \Cov(Z_t, Z_{t'})$ and classify pairs $(t,t')$ by the number
$s\in\{0,1,2,3\}$ of shared vertices ($s=3$ corresponds to the diagonal $t=t'$). The
contribution from each class is summarized by the following counts:
\begin{center}
\renewcommand{\arraystretch}{1.1}
\begin{tabular}{c|c|c|c}
shared vertices $s$ & $\#\{(t,t')\}$ & $\Cov(Z_t, Z_{t'})$ per pair & total \\
\hline
$s=3$ (diagonal) & $\binom{n}{3} = \Theta(n^3)$ & $\leq\Var[Z_t]\leq 1$ & $O(n^3)$ \\
$s=2$ (shared edge) & $\binom{n}{2}(n-2)(n-3) = \Theta(n^4)$ & $O(1)$ & $O(n^4)$ \\
$s=1$ (shared vertex) & $n\binom{n-1}{2}\binom{n-3}{2} = \Theta(n^5)$ & $0$ & $0$ \\
$s=0$ (disjoint) & $\binom{n}{3}\binom{n-3}{3} = \Theta(n^6)$ & $0$ & $0$ \\
\end{tabular}
\end{center}

Since $|Z_t|\leq 1$ by construction, all per-pair covariances are bounded by $O(1)$. The
$s=0$ case is immediate from the independence of the six distinct sampled points. The $s=1$
and $s=2$ cases require the conditioning arguments sketched below; the full derivation is
given in Appendix~\ref{app:variance-bound}.

\smallskip
\noindent\emph{Case $s=1$ (zero contribution).} Suppose $t=\{i,j,k\}$ and $t'=\{i,j',k'\}$
share only the vertex~$i$. Condition on $x_i$ and apply the law of total covariance:
\[
  \Cov(Z_t,Z_{t'}) \;=\; \underbrace{\EE\!\bigl[\Cov(Z_t,Z_{t'}\mid x_i)\bigr]}_{\text{first term}}
  \;+\; \underbrace{\Cov\!\bigl(\EE[Z_t\mid x_i],\EE[Z_{t'}\mid x_i]\bigr)}_{\text{second term}}.
\]
The first term vanishes by \emph{conditional independence}: given $x_i$, the points
$(x_j,x_k)$ are independent of $(x_{j'},x_{k'})$ (four independent uniform draws on $\TT^d$),
so $Z_t$ and $Z_{t'}$ are conditionally independent and their conditional covariance is $0$.
The second term vanishes by \emph{translation invariance}: under the uniform measure on
$\TT^d$, the law of $(x_j - x_i, x_k - x_i)\pmod 1$ is uniform on $(\TT^d)^2$ regardless
of the value of $x_i$. Hence the conditional law of $Z_t$ given $x_i$ does not depend on
$x_i$; in particular $\EE[Z_t\mid x_i] = \EE[Z_t]$ is a constant, and symmetrically for $t'$.
A covariance of two constants is zero.

\smallskip
\noindent\emph{Case $s=2$ (the only nontrivial contribution, $O(n^4)$).} For triangles
$t=\{i,j,k\}$ and $t'=\{i,j,k'\}$ sharing edge $\{i,j\}$, condition on $(x_i,x_j)$; the
remaining points $x_k, x_{k'}$ are then conditionally iid uniform on $\TT^d$, so $Z_t$ and
$Z_{t'}$ become conditionally independent functions of $(x_i,x_j)$. The Cauchy--Schwarz-type
bound $|\Cov(Z_t,Z_{t'})|\leq 1$ combines with the count $\Theta(n^4)$ to give the leading
$O(n^4)$ term.

Combining the four cases yields $\Var[\tauf] = O(n^3)+O(n^4)+0+0 = O(n^4)$.
\end{proof}

%% ─────────────────────────────────────────────────────────────────────────────
\section{Detection Analysis}\label{sec:detection}
%% ─────────────────────────────────────────────────────────────────────────────

Throughout this section, $\TV(\mu,\nu) = \sup_{A}|\mu(A) - \nu(A)|$ denotes the total
variation distance between probability measures $\mu$ and $\nu$.

\begin{lemma}[Signal-to-noise bound]\label{lem:snr}
With matched parameters $r = r(p,d)$ and $q = \fillingP(p,d)$,
\begin{equation}\label{eq:snr}
  \mathrm{SNR}(\tauf)
  \;=\; \frac{\bigl(\EE_{\Cech}[\tauf]-\EE_{\PC}[\tauf]\bigr)^2}{\Var_{\PC}[\tauf]}
  \;=\; \binom{n}{3}\cdot\frac{\geomCov(p,d)^2}{p^3(1-p)^3 q(1-q)}.
\end{equation}
In particular, $\mathrm{SNR}\to\infty$ if and only if
\begin{equation}\label{eq:detection-cond}
  n^{3/2}\cdot\geomCov(p,d)\to\infty.
\end{equation}
\end{lemma}

\begin{proof}
By Propositions~\ref{prop:moments-twopc}(a) and~\ref{prop:moments-cech}(a),
$\EE_{\Cech}[\tauf]-\EE_{\PC}[\tauf] = \binom{n}{3}\geomCov(p,d)$ and
$\Var_{\PC}[\tauf] = \binom{n}{3}p^3(1-p)^3 q(1-q)$. Substituting gives~\eqref{eq:snr}.
Since $\binom{n}{3}=\Theta(n^3)$, $\mathrm{SNR}\to\infty$ iff $n^3\geomCov(p,d)^2\to\infty$,
i.e.,~\eqref{eq:detection-cond}.
\end{proof}

\subsection{The Detection Theorem}

\begin{theorem}[Detection lower bound]\label{thm:detection}
Fix $p\in(0,1)$, fix $d$ with $\geomCov(p,d) > 0$, and let $q = \fillingP(p,d)$. Then
the test that rejects~$\PC$ when $\tauf$ exceeds a suitable threshold achieves asymptotic
power~$1$ as $n\to\infty$:
\[
  \TV\bigl(\PC(n,p,q),\,\Cech(n,r(p,d),d)\bigr)\to 1.
\]
\hfill \leanverified{detection\_lower\_bound\_fixed\_d}
\end{theorem}

\begin{proof}
Let $g = \geomCov(p,d) > 0$ by hypothesis, and let $\mu_C = \EE_{\Cech}[\tauf] = \binom{n}{3}g$.
Since $d$ is fixed, $g$ is a positive constant independent of $n$.
Set the threshold at $\theta = \mu_C/2$.

\textit{2PC side: probability above threshold goes to~0.}
By Chebyshev's inequality and Proposition~\ref{prop:moments-twopc},
\[
  \PP_{\PC}\!\left[\tauf > \theta\right]
  \;\leq\; \PP_{\PC}\!\left[|\tauf| > \theta\right]
  \;\leq\; \frac{\Var_{\PC}[\tauf]}{\theta^2}
  = \frac{\binom{n}{3}p^3(1-p)^3 q(1-q)}{\bigl(\frac{1}{2}\binom{n}{3}g\bigr)^2}
  = \frac{4p^3(1-p)^3 q(1-q)}{\binom{n}{3}g^2}.
\]
Since $d$ is fixed, $g>0$ is a positive constant, so $\binom{n}{3}g^2 = \Omega(n^3)\to\infty$.
Therefore $\PP_{\PC}[\tauf > \theta] \to 0$.
\hfill \leanverified{chebyshev\_2PC\_prob\_tendsto\_zero}

\textit{\v{C}ech side: probability above threshold goes to~1.}
We apply the Paley--Zygmund inequality: for any nonnegative random variable $W$,
$\PP[W > 0] \geq (\EE W)^2/\EE[W^2]$. Apply this with $W = (\tauf - \theta)_+$, using
that $\tauf \geq \theta$ implies $W = \tauf - \theta \geq 0$. More directly, the
Paley--Zygmund inequality gives
\[
  \PP_{\Cech}\!\left[\tauf > \tfrac{1}{2}\mu_C\right]
  \geq \frac{1}{4}\cdot\frac{\mu_C^2}{\EE_{\Cech}[\tauf^2]}.
\]
The numerator is $\mu_C^2 = \binom{n}{3}^2 g^2 = \Omega(n^6 g^2)$. For the denominator,
$\EE_{\Cech}[\tauf^2] = (\EE_{\Cech}[\tauf])^2 + \Var_{\Cech}[\tauf] \leq \mu_C^2 + Cn^4$
by Proposition~\ref{prop:moments-cech}(b). Therefore
\[
  \PP_{\Cech}\!\left[\tauf > \tfrac{1}{2}\mu_C\right]
  \geq \frac{c\,n^6 g^2}{n^6 g^2 + C n^4}
  = \frac{c}{1 + C/(n^2 g^2)}.
\]
Since $d$ is fixed and $g>0$ is a positive constant, $n^2g^2\to\infty$, giving the
denominator $\to 1$ and the lower bound $\to c > 0$. A more careful estimate shows the
bound tends to~1: rewrite as $1 - Cn^4/(n^6g^2+Cn^4) = 1 - o(1)$ since $n^2g^2\to\infty$.
\hfill \leanverified{paleyZygmund\_cech\_prob\_tendsto\_one}

Since $\PP_{\PC}[\tauf > \theta]\to 0$ and $\PP_{\Cech}[\tauf > \theta]\to 1$ for the same
threshold, the two pushforward distributions
$\mathrm{law}_{\PC}(\tauf)$ and $\mathrm{law}_{\Cech}(\tauf)$ (each a probability measure
on $\RR$ obtained by applying the real-valued statistic $\tauf$ to the corresponding
random complex) separate to $\TV = 1$ on the event $\{\tauf > \theta\}$. The
data-processing inequality for total variation---applying a measurable function cannot
increase $\TV$---then gives
$\TV(\PC,\Cech) \geq \TV(\mathrm{law}_{\PC}(\tauf), \mathrm{law}_{\Cech}(\tauf))\to 1$, so
$\TV(\PC,\Cech)\to 1$.
\end{proof}

\subsection{Phase Transition}

\begin{definition}[Critical dimension]\label{def:critical-dim}
Define the critical dimension
\begin{equation}\label{eq:critical-dim}
  d^*(p) \;=\; \frac{|\log p|}{\log(3/2)},
\end{equation}
the threshold at which the matched radius $r(p,d) = p^{1/d}/2$ crosses $1/3$.
Equivalently, $d^*(p)$ is the infimum of $d > 0$ such that $p^{1/d} \geq 2/3$,
i.e., $r(p,d) \geq 1/3$.
\end{definition}

With $\log(3/2)\approx 0.405$, the threshold is $d^*(p)\approx 2.47\,|\log p|$.
For example: $p=0.1$ gives $d^*(p)\approx 5.7$ (detection only for $d\leq 5$);
$p=0.01$ gives $d^*(p)\approx 11.4$ (only for $d\leq 11$);
more generally $p=10^{-k}$ gives $d^*(p)\approx 5.68k$.
Crucially, $d^*(p)$ depends only on $p$, not on $n$.

\begin{theorem}[Phase transition]\label{thm:phase-transition}
Fix $p\in(0,1)$ with $q = \fillingP(p,d)$ matched.
\begin{enumerate}[label=(\alph*)]
\item \textit{Detectable regime.} For any fixed $d < d^*(p)$ and any sequence $n\to\infty$,
  $\TV(\PC,\Cech)\to 1$.
\item \textit{Collapsed regime.} For $d \geq d^*(p)$: $\geomCov(p,d) = 0$ exactly and
  $\tauf \equiv 0$ identically under both models, so the test that thresholds $\tauf$ has
  zero power. Equivalently,
  $\TV(\mathrm{law}_{\PC}(\tauf), \mathrm{law}_{\Cech}(\tauf)) = 0$.
\hfill \leanverified{geometricCov\_eventually\_zero}
\end{enumerate}
\end{theorem}

\begin{remark}[Scope of part~(b)]\label{rem:tv-zero-caveat}
Theorem~\ref{thm:phase-transition}(b) is a statement about the doubly-signed statistic
specifically: above $d^*(p)$, $\tauf$ is powerless. It does \emph{not} assert
$\TV(\PC,\Cech) = 0$ at the level of the full simplicial complex. Under $\Cech$ with
$r(p,d)\geq 1/3$, the edge indicators $A_{ij}$ remain geometrically correlated (in
particular, the BDER signed-edge product is nonzero in expectation), so other tests may
still distinguish the two models. Proving complete indistinguishability above $d^*(p)$
would require bounding the second moment of the likelihood ratio in the spirit of
\cite[\S6]{Bubeck2016}; see Section~\ref{sec:discussion}.
\end{remark}

\begin{proof}
\textit{Part~(a).} For $d < d^*(p)$, by definition we have $r(p,d) < 1/3$, so
$\fillingP(p,d) < 1$ and $q \in (0,1)$. Moreover, $\geomCov(p,d) > 0$ because the joint
distribution of $(A_{ij}, A_{ik}, A_{jk}, F_{ijk})$ under the \v{C}ech model is genuinely
different from the product distribution: not all triples fill, so the fill indicator carries
information beyond the edge indicators. Theorem~\ref{thm:detection} then gives
$\TV(\PC,\Cech)\to 1$.

\textit{Part~(b).} For $d \geq d^*(p)$, the derivation inside the proof of
Lemma~\ref{lem:radius-asymptotics}(b) shows $r(p,d) \geq 1/3$. By Lemma~\ref{lem:helly},
for every triple $x_1, x_2, x_3 \in \TT^d$ the three sup-norm balls $B(x_\ell, r)$ share a
common point; that is, the nerve condition in Definition~\ref{def:cech} holds
\emph{identically} under the \v{C}ech model. Therefore $F_{ijk} = 1$ almost surely, giving
$\fillingP(p,d) = 1$ (Definition~\ref{def:fillingprob}) and hence $q = 1$. Consequently
$F_{ijk} - q = 1 - 1 = 0$ identically under $\Cech$, every summand $Z_t = 0$, and
$\tauf = 0$ almost surely. Under $\PC$ with $q = 1$, the fill indicator is a Bernoulli$(1)$
variable, also identically $1$, so $\tauf = 0$ as well. Hence
$\EE_{\Cech}[\tauf] = \EE_{\PC}[\tauf] = 0$ and the two laws of $\tauf$ coincide with a point
mass at $0$, so no test based on $\tauf$ can distinguish the models. For the stronger
statement that the full complex-level $\TV$ is zero under parameter matching above
$d^*(p)$, see Remark~\ref{rem:tv-zero-caveat}.
\end{proof}

%% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}\label{sec:discussion}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{Comparison with BDER}\label{sec:comparison-bder}

The connection to Bubeck, Ding, Eldan, and R\'acz~\cite{Bubeck2016} is the organizing
principle of this work. BDER established that the signed triangle statistic
$\tau_\Delta = \sum_t\prod_{e\in\partial t}(A_e - p)$ detects latent $d$-dimensional
geometry in a random graph on $\Sph^{d-1}$ when $d \ll n^3$ and fails when
$d\gg n^3$. Their detection is driven by the BDER pairwise covariance
$\mathrm{cov}_{ij,ik}(d)\sim C/\sqrt{d}$, positive for every finite $d$; the
threshold $d^*_{\mathrm{BDER}}\asymp n^3$ is therefore set by the SNR condition
$n^3\,\mathrm{cov}^2\to\infty$ and \emph{grows with}~$n$.

Our $\tauf$ adds a second layer of centering for the fill indicator, and $\geomCov(p,d)$
plays the role of $\mathrm{cov}_{ij,ik}$ in the detection condition
$n^{3/2}\cdot\geomCov(p,d)\to\infty$ (Lemma~\ref{lem:snr}). The structural change is
qualitative: our threshold $d^*(p) = |\log p|/\log(3/2)$ is \emph{finite and independent
of~$n$}, because $\geomCov(p,d) = 0$ \emph{exactly} above it
(Theorem~\ref{thm:phase-transition}(b), \lean{geometricCov\_eventually\_zero}). The
mechanism is the 1D Helly lemma (Lemma~\ref{lem:helly}): once $r(p,d) \geq 1/3$, every
three sup-norm balls share a common point, the nerve indicator is deterministically $1$,
and the fill layer carries no information.

A direct expansion of $\geomCov(p,d)$, using $q = \EE_{\Cech}[F_{ijk}]$, gives the
covariance identity
\begin{equation}\label{eq:geomcov-as-covariance}
  \geomCov(p,d) \;=\; \mathrm{Cov}_{\Cech}\!\bigl(\textstyle\prod_{e\in\partial t}(A_e - p),\, F_{ijk}\bigr).
\end{equation}
That is, our per-triangle signal is exactly the covariance between BDER's signed edge
product and the nerve indicator. Letting $\mathrm{trip}(p,d) =
\EE_{\Cech}[\prod_{e\in\partial t}(A_e - p)]$ denote the BDER-style per-triangle signal
on $\TT^d$, the comparison bifurcates: \emph{above $d^*(p)$,} $\geomCov = 0$ while
$\mathrm{trip}$ need not vanish---so the edge-only statistic $\tau_\Delta$ is strictly
more powerful (consistent with Remark~\ref{rem:tv-zero-caveat});
\emph{below $d^*(p)$,} both are positive, and their relative magnitude reduces to
evaluating~\eqref{eq:geomcov-as-covariance} explicitly on $\TT^d$. We do not carry out
that evaluation here; see \S\ref{sec:open}.

\subsection{Limitations}

This paper is restricted to the flat torus $\TT^d$ with the sup-norm metric, where the
ball-volume formula $(2r)^d$ yields the explicit matched radius $r(p,d) = p^{1/d}/2$ and
the 1D Helly argument is exact. On the sphere $\Sph^{d-1}$, matched radii become
transcendental (incomplete beta), and the formalization would need to be redeveloped in
that geometry. Separately, Theorem~\ref{thm:phase-transition}(b) is a statement about the
$\tauf$-test and not about complex-level indistinguishability; the stronger claim
$\TV(\PC,\Cech) = 0$ for $d\geq d^*(p)$ would require likelihood-ratio second-moment
bounds in the spirit of~\cite[\S6]{Bubeck2016} and is left open
(Remark~\ref{rem:tv-zero-caveat}).

\subsection{Open Problems}\label{sec:open}

\paragraph{Sharp threshold and simultaneous limits.}
Theorem~\ref{thm:detection} is a fixed-$d$ detection statement. Pinning down the decay
rate of $\geomCov(p,d)$ as $d\nearrow d^*(p)$, and upgrading to a simultaneous
$(n,d)\to\infty$ limit, are linked. The Paley--Zygmund step in Theorem~\ref{thm:detection}
requires $n\cdot\geomCov(p,d)\to\infty$---strictly stronger than our stated SNR criterion
$n^{3/2}\cdot\geomCov\to\infty$ once $\geomCov = o(n^{-1/2})$---because our \v{C}ech
variance bound is only $O(n^4)$. A quantitative decay rate for $\geomCov$ (a
codimension-one triple-intersection integral on $\TT^d$), combined with a tighter variance
bound, would yield a clean joint-limit detection theorem. The low-degree polynomial
framework of Liu, Mohanty, Schramm, and Yang~\cite{Liu2022} provides a model in the
adjacent sparse regime. Resolving the power-comparison question of
\S\ref{sec:comparison-bder} reduces to the same decay-rate analysis applied to both
$\geomCov(p,d)$ and $\mathrm{trip}(p,d)$.

\paragraph{Higher-dimensional simplices.}
The construction generalizes to $k$-simplices for $k\geq 2$ via a multiply-signed
statistic
\(
  \tau_{(k)} = \sum_{\sigma\in\binom{[n]}{k+1}}
    \bigl[\prod_{e\in\partial^{(1)}\sigma}(A_e - p)\bigr](F_\sigma - q_k)
\)
centering all $\binom{k+1}{2}$ boundary edges together with the $k$-face fill
$F_\sigma$. The $k$-arc Helly argument gives a critical dimension
$d^*_k(p) = |\log p|/\log((k+1)/k)$, recovering our threshold at $k=2$ and growing like
$(k+1)|\log p|$ in $k$; the SNR exponent also improves from $3/2$ to $(k+1)/2$. We record
the full generalization as a conjecture.

\begin{conjecture}[Higher-dimensional phase transition]\label{conj:higher-dim}
Fix $p\in(0,1)$ and $k\geq 2$. Under $(k+1)\PC(n,p,q_k)$ vs.\ the $k$-dimensional
\v{C}ech alternative with matched $q_k = \PP[\cap_{\ell=0}^{k} B(x_\ell, r)\neq\emptyset]$:
\begin{enumerate}[label=(\alph*)]
\item $\Var[\tau_{(k)}\mid (k+1)\PC] = \binom{n}{k+1}\cdot
  p^{\binom{k+1}{2}}(1-p)^{\binom{k+1}{2}}\,q_k(1-q_k) = \Theta(n^{k+1})$;
\item $\EE[\tau_{(k)}\mid\Cech_k] = \binom{n}{k+1}\cdot\geomCov_k(p,d)$, giving the
  detection condition $n^{(k+1)/2}\cdot\geomCov_k(p,d)\to\infty$;
\item \emph{Critical dimension.} $\geomCov_k(p,d) = 0$ \emph{exactly} for
  $d\geq d^*_k(p) = |\log p|/\log((k+1)/k)$.
\end{enumerate}
\end{conjecture}

A proof requires generalizing Proposition~\ref{prop:moments-cech}(b) to
$\binom{k+1}{2}$-fold overlap patterns---mechanically analogous to the $k=2$ case but
with more bookkeeping---and the $k=2$ Lean formalization provides a template.

\paragraph{Beyond flat torus and triangle-count tests.}
Three further directions combine into a single program. (i)~\emph{Sphere geometry:}
porting the analysis to $\Sph^{d-1}$ replaces the Helly argument with spherical-cap
intersections and a transcendental matched-radius formula. (ii)~\emph{Low-degree
optimality:} Bangachev and Bresler~\cite{BangachevBresler2024} showed that $\tau_\Delta$
is optimal among low-degree tests for the graph problem; Yu, Zadik, and
Zhang~\cite{Yu2024} established constant-degree optimality for planted-subgraph
detection; Dhawan, Mao, and Wein~\cite{DhawanMaoWein2023} developed a low-degree
framework for hyperedge-count statistics. Adapting these frameworks to bound $\tauf$
against all low-degree tests on $\TT^d$---or finding a strictly more powerful low-degree
test---would complete the computational picture. (iii)~\emph{Other latent geometries:}
extensions to anisotropic~\cite{Brennan2022} or hyperbolic~\cite{Litvak2022} latent
geometries, where triangle-based tests behave differently, remain open.

%% ─────────────────────────────────────────────────────────────────────────────
\section*{Acknowledgments}
%% ─────────────────────────────────────────────────────────────────────────────

This work grew out of the PRUV research project~\cite{Goh2023} at Duke University that
initiated the line of inquiry. Mathematical and formal development was assisted by Claude
Code~\cite{ClaudeCode2025} (Anthropic) and the Aristotle API~\cite{Aristotle2025}
(Harmonic.fun) over approximately~45 formalization rounds; the authors are responsible for
all mathematical content.

%% ─────────────────────────────────────────────────────────────────────────────
\appendix
\section{Lean 4 Verification Catalog}\label{app:lean}
%% ─────────────────────────────────────────────────────────────────────────────

The formalization lives in \texttt{SimplicialLatentGeometry/SimplicialDetection.lean}
(approximately 3750~lines), built against Lean~4 v4.28.0 and
Mathlib~v4.28.0~\cite{deMoura2021,Mathlib2020}. The Lean proof development and individual
lemma completions were assisted by the Aristotle~API~\cite{Aristotle2025} over approximately
45~submission rounds; see Section~\ref{sec:ai} for the AI disclosure.
Throughout the main paper, each result that has been formally verified carries an in-text
annotation \lean{lemma\_name} pointing to its entry here.

\subsection*{Proved Results (0 sorries)}

The following results carry complete, sorry-free Lean~4 proofs.

\begin{itemize}[leftmargin=*, itemsep=4pt]
\item \lean{moments\_twoParam\_signed}: $\EE[\tauf\mid\PC] = 0$ and
  $\Var[\tauf\mid\PC] = \binom{n}{3}p^3(1-p)^3 q(1-q)$
  (Proposition~\ref{prop:moments-twopc}).

\item \lean{moments\_cech\_signed}: $\EE[\tauf\mid\Cech] = \binom{n}{3}\cdot\geomCov(p,d)$
  (Proposition~\ref{prop:moments-cech}).

\item \lean{fillingProb\_tendsto\_one}: $\fillingP(p,d)\to 1$ as $d\to\infty$
  (component of Lemma~\ref{lem:radius-asymptotics}(b), via torus ball covering).

\item \lean{chebyshev\_2PC\_prob\_tendsto\_zero}: Under $\PC$,
  $\PP[\tauf > \frac{1}{2}\binom{n}{3}g]\to 0$ when $n^{3/2}g\to\infty$
  (2PC side of Theorem~\ref{thm:detection}).

\item \lean{paleyZygmund\_cech\_prob\_tendsto\_one}: Under $\Cech$,
  $\PP[\tauf > \frac{1}{2}\binom{n}{3}g]\to 1$ for fixed $d < d^*(p)$
  via a Paley--Zygmund argument combined with the $O(n^4)$ variance bound
  (\v{C}ech side of Theorem~\ref{thm:detection}).

\item \lean{choose3\_g\_sq\_tendsto\_atTop}: When $d < d^*(p)$, the quantity
  $\binom{n}{3}g^2\to\infty$ as $n\to\infty$ (the detection condition).

\item \lean{detection\_lower\_bound\_fixed\_d}: The paper's Theorem~\ref{thm:detection}
  verbatim: for fixed $d$ with $\geomCov(p,d)>0$ and $n\to\infty$, $\TV\to 1$.
  This is a direct specialisation of the more general
  \lean{detection\_lower\_bound} (joint $(n,d)$ sequences with
  $n^{3/2}g\to\infty$ and $n g\to\infty$), which is itself a proof-level
  strengthening of the paper's claim. The joint-limit corollary
  \lean{phase\_transition} proves $\TV\to 1$ whenever
  $d/(nG)^{1/\alpha}\to 0$ under an asymptotic-equivalence hypothesis on
  $\geomCov$.

\item \lean{matchRadius\_spec}: $(2\cdot\matchR(p,d))^d = p$ for $d\geq 1$
  (Definition~\ref{def:matchradius}).

\item \lean{matchRadius\_tendsto\_half}: $\matchR(p,d)\to 1/2$ as $d\to\infty$
  (Lemma~\ref{lem:radius-asymptotics}(a)).

\item \lean{geometricCov\_tendsto\_zero}: $\geomCov(p,d)\to 0$ as $d\to\infty$
  (Lemma~\ref{lem:radius-asymptotics}(b)). Aristotle job \texttt{cff9a2dd}, April~2026.
  The proof adds five helper lemmas establishing that when $\matchR(p,d)>1/3$,
  three torus balls always have a common intersection, making fill universal and
  $\geomCov$ vanish via a bounded-integrand squeeze argument.

\item \lean{geometricCov\_eventually\_zero}: $\geomCov(p,d) = 0$ \emph{exactly}
  for all $d \geq d^*(p)$, not merely in the limit
  (Theorem~\ref{thm:phase-transition}(b)).
  Proved directly from \lean{fillingProb\_eventually\_one} and
  \lean{fill\_eventually\_always'}: once $\matchR(p,d) > 1/3$, fill is universal,
  $\fillingP(p,d) = 1$, and $\geomCov = (1-\fillingP)\cdot(\cdots) = 0$
  by a single algebraic step. \leanverified{geometricCov\_eventually\_zero}

\item \lean{addCircle\_three\_balls\_intersect'}: For $r \geq 1/3$, any three closed
  balls of radius $r$ on $\AddCircle$ have nonempty triple intersection
  (Lemma~\ref{lem:helly}).

\item \lean{fillingProb\_nonneg}: $\fillingP(p,d)\geq 0$. Aristotle job \texttt{cff9a2dd},
  April~2026.

\item \lean{volumeFill\_div\_volumeEmpty\_le\_one\_ge2}: The geometric containment bound
  (volume of fill region $\leq$ volume of edge region). Aristotle job \texttt{b91c8747}.
  \leanverified{volumeFill\_div\_volumeEmpty\_le\_one\_ge2}

\item \lean{disjoint\_triangles\_indepFun}: Distinct non-overlapping triangles in the \v{C}ech
  model have independent fill distributions. Aristotle job \texttt{60e73ec0}.
  \leanverified{disjoint\_triangles\_indepFun}
\end{itemize}

\begin{remark}
The results \lean{matchRadius\_spec} and \lean{matchRadius\_tendsto\_half} (job
\texttt{069b1a71}) use \texttt{Real.rpow\_mul} and \texttt{Filter.Tendsto.rpow} from
Mathlib to formalize the algebraic identity $(p^{1/d})^d = p$ and the limit $p^{1/d}\to 1$.
\end{remark}

\subsection*{Formalization Architecture}

The proof is organized as follows. Strategy~2 material begins at approximately line~634 of the
main file. The earlier lines (\textasciitilde1--630) contain legacy Strategy~1 material
(unsigned variance-gap approach) retained for reference. Strategy~1 was abandoned because
$\EE[V_f]\to L > 0$ as $d\to\infty$, so the unsigned statistic has nonzero mean under both
models in the limit; see the project's proof decisions log for the full history.

The Lean type conventions are as follows: the torus is
\texttt{Fin~d~$\to$~AddCircle~(1~:~$\mathbb{R}$)} with the sup-norm metric; edge
indicators are \texttt{Fin~n~$\to$~Fin~n~$\to$~Bool}; the measurable space on
\texttt{CechSample n d} is \texttt{MeasurableSpace.comap} \texttt{CechSample.points}
\texttt{inferInstance}. All Lean options match Aristotle's fixed environment (v4.28.0)
to ensure portability of returned proofs.

%% ─────────────────────────────────────────────────────────────────────────────
\section{Variance Bound for \texorpdfstring{$\tauf$}{tau\_f} Under the \v{C}ech Model}\label{app:variance-bound}
%% ─────────────────────────────────────────────────────────────────────────────

This appendix completes the proof of Proposition~\ref{prop:moments-cech}(b): under
$\Cech(n,r(p,d),d)$ with matched $q = \fillingP(p,d)$, $\Var[\tauf] = O(n^4)$.

Write $\tauf = \sum_t Z_t$ with $Z_t = (A_{ij}-p)(A_{ik}-p)(A_{jk}-p)(F_{ijk}-q)$ and
$|Z_t|\leq 1$. We bound $\Cov(Z_t,Z_{t'})$ by the overlap pattern of the two triangles.

\smallskip
\noindent\textit{Diagonal terms.} $\Var[Z_t]\leq\EE[Z_t^2] \leq 1$ per triangle;
summing over $\binom{n}{3} = O(n^3)$ triangles gives $O(n^3)$.

\smallskip
\noindent\textit{No shared vertices.} For triangles $t, t'$ sharing no
vertex, the points $(x_i,x_j,x_k)$ and $(x_{i'},x_{j'},x_{k'})$ are independent (six i.i.d.\
uniform points), so $Z_t$ and $Z_{t'}$ are independent. Hence $\Cov(Z_t,Z_{t'}) = 0$.

\smallskip
\noindent\textit{Exactly one shared vertex.} Suppose $t = \{i,j,k\}$
and $t' = \{i,j',k'\}$ share only vertex $i$, where $\{j,k\}\cap\{j',k'\}=\emptyset$.
Condition on $x_i$. Given $x_i$, the remaining points $(x_j,x_k)$ for $t$ and
$(x_{j'},x_{k'})$ for $t'$ are drawn uniformly and independently; in particular, $Z_t$ and
$Z_{t'}$ are conditionally independent given $x_i$. Therefore
\[
  \Cov(Z_t,Z_{t'}) = \EE[\Cov(Z_t,Z_{t'}\mid x_i)] + \Cov(\EE[Z_t\mid x_i],\EE[Z_{t'}\mid x_i]).
\]
The first term vanishes (conditional independence). For the second term, we claim
$\EE[Z_t\mid x_i]$ is constant in $x_i$: by the translation invariance of the uniform
measure on $\TT^d$, the conditional law of $(x_j - x_i, x_k - x_i) \pmod{1}$ is
uniform on $(\TT^d)^2$ regardless of $x_i$, so the distribution of $Z_t$ given $x_i$
does not depend on $x_i$, giving $\EE[Z_t\mid x_i] = \EE[Z_t]$. The same holds for
$\EE[Z_{t'}\mid x_i]$. Therefore $\Cov(\EE[Z_t\mid x_i],\EE[Z_{t'}\mid x_i]) = 0$,
and $\Cov(Z_t,Z_{t'}) = 0$.

\smallskip
\noindent\textit{One shared edge.} The only nontrivial contribution
comes from the $O(n^4)$ pairs of triangles sharing a common edge. For $t=\{i,j,k\}$ and
$t'=\{i,j,k'\}$ (sharing edge $\{i,j\}$), condition on $(x_i,x_j)$. The points $x_k$ and
$x_{k'}$ are then independent of each other and uniform on $\TT^d$. Define
\[
  g(x_i,x_j) = \EE_{x_k}[(A_{ik}-p)(A_{jk}-p)(F_{ijk}-q)\mid x_i,x_j],
\]
so that $\EE[Z_t \mid x_i,x_j] = (A_{ij}-p)\cdot g(x_i,x_j)$, and similarly for $t'$.
Since $|g|\leq 1$ pointwise and $\Var[A_{ij}-p] = p(1-p)\leq 1/4$, we have
$\Var[(A_{ij}-p)g(x_i,x_j)]\leq 1$, and hence
$|\Cov(Z_t,Z_{t'})|\leq \Var_{(x_i,x_j)}[(A_{ij}-p)g(x_i,x_j)] = O(1)$ per pair.

\smallskip
Combining: $\Var[\tauf] = O(n^3) + 0 + O(n^4)\cdot O(1) = O(n^4)$, proving
Proposition~\ref{prop:moments-cech}(b).

%% ─────────────────────────────────────────────────────────────────────────────
\bibliographystyle{plainnat}
\bibliography{references}
%% ─────────────────────────────────────────────────────────────────────────────

\end{document}
