%% paper.tex — Testing for Simplicial Geometry
%% arXiv preprint targeting Random Structures and Algorithms (Wiley)
%% Compile: cd simplicial-latent-geometry/my_theorems && pdflatex paper && bibtex paper && pdflatex paper && pdflatex paper

\documentclass[reqno,12pt]{amsart}

%% ── Encoding & fonts ─────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}

%% ── Mathematics ──────────────────────────────────────────────────────────────
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}          % \coloneqq, dcases

%% ── Bibliography ─────────────────────────────────────────────────────────────
\usepackage[numbers,sort&compress]{natbib}

%% ── Hyperlinks ───────────────────────────────────────────────────────────────
\usepackage[dvipsnames,svgnames,x11names]{xcolor}
\usepackage[colorlinks=true,
            linkcolor=RoyalBlue4,
            citecolor=RoyalBlue4,
            urlcolor=NavyBlue]{hyperref}
\usepackage{url}

%% ── Page geometry ────────────────────────────────────────────────────────────
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}

%% ── Enumitem for tighter lists ───────────────────────────────────────────────
\usepackage{enumitem}

%% ─────────────────────────────────────────────────────────────────────────────
%%  MACROS
%% ─────────────────────────────────────────────────────────────────────────────

%% Models and statistics
\newcommand{\tauf}{\tau_{\mathrm{f}}}
\newcommand{\PC}{\mathrm{2PC}}
\newcommand{\Cech}{\ifmmode\text{\v{C}ech}\else\v{C}ech\fi}  % Čech (text+math)
\newcommand{\RGG}{\mathrm{RGG}}
\newcommand{\TV}{\mathrm{TV}}

%% Named quantities
\newcommand{\geomCov}{\mathrm{geomCov}}
\newcommand{\fillingP}{\mathrm{fillingProb}}
\newcommand{\matchR}{\mathrm{matchRadius}}

%% Sets and spaces
\newcommand{\TT}{\mathbb{T}}         % flat torus
\newcommand{\AddCircle}{\mathbb{T}^1}% AddCircle(1) = ℝ/ℤ, the unit circle
\newcommand{\RR}{\mathbb{R}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\Sph}{\mathbb{S}}

%% Probability
\newcommand{\EE}{\mathbb{E}}
\newcommand{\PP}{\mathbb{P}}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}

%% Lean annotation (gray box for theorem names)
\newcommand{\lean}[1]{\textup{\texttt{\small #1}}}
\newcommand{\leanverified}[1]{\textup{(Lean: \texttt{#1}, 0 sorries)}}


%% ─────────────────────────────────────────────────────────────────────────────
%%  THEOREM ENVIRONMENTS
%% ─────────────────────────────────────────────────────────────────────────────

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{observation}[theorem]{Observation}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{conjecture}[theorem]{Conjecture}

%% ─────────────────────────────────────────────────────────────────────────────
%%  DOCUMENT
%% ─────────────────────────────────────────────────────────────────────────────

\begin{document}

\title[Testing for Simplicial Geometry]{Testing for Simplicial Geometry: A Signed Triangle
  Statistic for the \v{C}ech--2PC Phase Transition}

\author{David Goh}
\address{Department of Computer Science, University of Toronto}
\email{daveed@cs.toronto.edu}

\author{Nicholas A.\ Cook}
\address{Department of Mathematics, Duke University}

\date{}

\subjclass[2020]{%
  Primary 05C80 (random graphs, generic models);
  Secondary 55U10 (simplicial sets and complexes),
  62H30 (classification and discrimination; cluster analysis).}

\keywords{random simplicial complex, geometry detection, Čech complex, signed triangle statistic,
  phase transition, Lean 4, formal verification}

\begin{abstract}
Random simplicial complexes are natural higher-dimensional analogues of random graphs, arising
in topological data analysis and the study of networks with higher-order structure. A basic
question in statistical inference on such objects is whether an observed complex reflects
latent geometric organization---in which the vertices have positions in some ambient space
that drive the connection structure---or whether its triangle structure appeared by chance,
with no underlying geometry. We study a precise formulation of this geometry detection
problem: given an observed 2-dimensional simplicial complex on $n$ vertices, determine whether
it arose from the \emph{2-parameter complex} $\PC(n,p,q)$---in which edges and triangle fills
are independent---or from the \v{C}ech complex of $n$ points sampled uniformly on the flat
torus $\TT^d$, with radius chosen to match the edge probability~$p$.

We introduce the \emph{doubly-signed filled triangle statistic} $\tauf$, centered with respect
to both edge and fill indicators. Under $\PC$, $\tauf$ has mean zero and variance $\Theta(n^3)$.
Under the \v{C}ech model, $\tauf$ has mean $\binom{n}{3}\cdot\geomCov(p,d)$, where
$\geomCov(p,d)\to 0$ as $d\to\infty$. There is an explicit sharp threshold
$d^*(p) = |\log p|/\log(3/2)$: for $d \geq d^*(p)$, every triple of points on $\TT^d$ fills
and $\geomCov(p,d) = 0$ \emph{exactly}, so detection is impossible; for $d < d^*(p)$,
$\geomCov(p,d) > 0$ and detection succeeds when $n^{3/2}\cdot\geomCov(p,d)\to\infty$.
All moment results, detection lower bound, the phase transition $\geomCov(p,d)=0$ for $d\geq d^*(p)$,
and the limit $\geomCov(p,d)\to 0$ are formally verified in Lean~4 with
Mathlib~\cite{Mathlib2020}, with no remaining sorries in the active proof chain.
\end{abstract}

\maketitle

%% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{The Detection Problem}

Networks in the real world often exhibit higher-order structure: not just pairwise connections
between entities, but groups of three or more that are jointly related. Simplicial complexes
formalize this by augmenting a graph with filled triangles, tetrahedra, and higher faces.
Random simplicial complexes serve as null models for such data, and a natural question arises:
\emph{does an observed complex carry latent geometric structure, or did it arise by a purely
combinatorial chance process?}

To make this precise, we focus on 2-dimensional random simplicial complexes---those with
vertices, edges, and filled triangles. The simplest null model is the
\emph{2-parameter complex} $\PC(n,p,q)$: place $n$ vertices, include each edge independently
with probability $p$, and fill each completed triangle independently with probability $q$.
This is a maximally unstructured model---edge and fill events are all mutually independent,
encoding no geometry whatsoever.

The alternative is a \emph{geometric} model. Sample $n$ points uniformly on the flat torus
$\TT^d$; connect two points by an edge whenever they lie within distance $r$; fill a triangle
whenever all three associated balls of radius $r$ have a common point (the nerve criterion).
This is the \v{C}ech complex $\Cech(n,r,d)$, and its triangle structure is governed by the
ambient positions of the points, not by independent coin flips. The detection question is
whether, given only the simplicial complex, one can tell which model generated it.

This type of question has been studied in the graph setting by Bubeck, Ding, Eldan, and
R\'acz~\cite{Bubeck2016}, who established a sharp phase transition: random geometric graphs
on $\Sph^{d-1}$ are distinguishable from Erd\H{o}s--R\'enyi graphs $G(n,p)$ when
$d = O(n^3/\log n)$, and indistinguishable when $d\gg n^3$. Their key tool was a signed
triangle statistic that exploits the fact that edges sharing a vertex are positively correlated
under the geometric model. Our work carries this investigation into the simplicial setting,
where the presence of fill indicators introduces an additional layer of geometric dependence
beyond what edge structure alone can capture.

\subsection{Main Contributions}

\paragraph{Core idea.}
The key insight is a two-layer centering. Under both models, each edge has the same marginal
probability $p$ and each fill has the same marginal probability $q$ (by parameter matching),
so no single indicator separates them. What distinguishes the \v{C}ech model is
\emph{joint} dependence: the three edges and the fill of a triangle are all functions of the
same three ambient points, making them collectively dependent in a way that the independent
2PC cannot replicate. The doubly-signed statistic $\tauf$ isolates this joint dependence by
centering both the edge product and the fill indicator simultaneously---under 2PC, the
product has expectation zero by independence; under \v{C}ech, geometric correlations make it
nonzero. As dimension grows, the matched ball radius approaches $1/2$ (the torus diameter),
balls eventually cover the whole torus, fill becomes trivially universal, and the geometric
covariance vanishes---giving the phase transition.

\begin{theorem}[Informal]\label{thm:informal}
Fix $p\in(0,1)$. There is a critical dimension
$d^*(p) = |\log p|/\log(3/2) \approx 2.47\,|\log p|$ such that:
\begin{itemize}
  \item \textbf{Detectable:} For $d < d^*(p)$ and $n^{3/2}\cdot\geomCov(p,d)\to\infty$,
    the doubly-signed test $\tauf$ achieves asymptotic power~$1$.
  \item \textbf{Collapsed:} For $d \geq d^*(p)$, $\geomCov(p,d) = 0$ exactly and
    $\tauf$ is identically distributed under $\PC$ and $\Cech$---no test based on
    $\tauf$ can distinguish the models.
\end{itemize}
\end{theorem}

We make the following contributions.

\begin{enumerate}[label=(\arabic*)]
\item \textbf{The doubly-signed statistic.} We define
  \begin{equation}\label{eq:tauf}
    \tauf = \sum_{\{i,j,k\}\in\binom{[n]}{3}}
      \Bigl[\prod_{e\in\partial\{i,j,k\}}(A_e - p)\Bigr](F_{ijk} - q),
  \end{equation}
  a statistic centered with respect to both edge and fill variability. This is the natural
  simplicial analogue of the BDER signed triangle statistic.

\item \textbf{Moment computations.} We prove that under $\PC(n,p,q)$,
  $\EE[\tauf] = 0$ and $\Var[\tauf] = \binom{n}{3}p^3(1-p)^3 q(1-q)$. Under
  $\Cech(n,r,d)$ with matched parameters, $\EE[\tauf] = \binom{n}{3}\cdot\geomCov(p,d)$.

\item \textbf{Phase transition.} We identify the explicit critical dimension
  $d^*(p) = |\log p|/\log(3/2)$: for $d \geq d^*(p)$ the matched ball radius exceeds $1/3$,
  every triple fills by a 1D Helly argument, and $\geomCov(p,d) = 0$ exactly
  (not merely small). For $d < d^*(p)$, the SNR condition $n^{3/2}\cdot\geomCov(p,d)\to\infty$
  drives detection.

\item \textbf{Lean verification.} The moment results, detection lower bound, phase
  transition (both regimes), and $\geomCov(p,d) = 0$ for $d \geq d^*(p)$ are formally
  verified in Lean~4 with Mathlib, with no remaining sorries in the active proof chain.
\end{enumerate}

\subsection{Comparison with BDER}

The BDER problem compares $G(n,p)$ with the random geometric graph $\RGG(n,p,d)$ on
$\Sph^{d-1}$. Their signed statistic exploits the fact that edge indicators in a triangle
are positively correlated under the geometric model (sharing a common ambient point), so
$\EE[\prod(A_e-p)] > 0$ under $\RGG$ but $= 0$ under $G(n,p)$.

In our setting, the analogous structure requires a second layer of centering. Even if the
edge indicators are correlated, the fill indicator $F_{ijk}$ carries additional geometric
information: under the \v{C}ech model, the fill event depends on all three points
simultaneously, not just on pairwise distances. The doubly-signed statistic $\tauf$ captures
precisely this joint dependence. The geometric covariance $\geomCov(p,d)$ is the simplicial
analogue of the BDER covariance $\mathrm{cov}_{ij,ik}$.

The detection threshold in BDER is $d^*\asymp n^3$ (growing with $n$). Our threshold
$d^*(p) = |\log p|/\log(3/2)$ is \emph{finite and independent of $n$}: above it,
$\geomCov(p,d) = 0$ exactly and detection is impossible regardless of sample size.
This sharp collapse---rather than a gradual SNR decay---is the distinctive feature of
the simplicial setting; see Section~\ref{sec:threshold} for the geometric derivation.

\subsection{Paper Organization}

Section~\ref{sec:models} defines the models and parameter matching.
Section~\ref{sec:statistic} introduces $\tauf$ and computes its moments under both models.
Section~\ref{sec:detection} carries out the detection analysis and states the main theorems.
Section~\ref{sec:lean} describes the Lean~4 formalization.
Section~\ref{sec:discussion} discusses implications and open problems.

\subsection{AI-Assisted Mathematical Development}\label{sec:ai}

The development of this paper was assisted by two AI tools.
\textbf{Claude Code}~\cite{ClaudeCode2025} (Anthropic) contributed at both the mathematical
and formal levels: it assisted with the formulation of the doubly-signed statistic~$\tauf$,
the BDER analogy as a guiding framework, and the phase transition analysis, as well as proof
architecture, tactic selection, and Lean~4 syntax over approximately~45 formalization rounds.
The \textbf{Aristotle API}~\cite{Aristotle2025} (Harmonic.fun) performed automated formal
proof completion for individual lemmas; each returned proof was reviewed and cherry-picked
into the codebase by the authors.

This work follows the approach of Jang and Ryu~\cite{JangRyu2026}, who employed ChatGPT in
the \emph{discovery} of their proof of point convergence of Nesterov's accelerated gradient
method. As in that work, AI assistance extended beyond the formal verification layer into
mathematical exploration. The authors are responsible for all mathematical content.

%% ─────────────────────────────────────────────────────────────────────────────
\section{Models and Parameter Matching}\label{sec:models}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{The 2-Parameter Complex}

\begin{definition}[2-parameter complex]
The \emph{2-parameter complex} $\PC(n,p,q)$ is the random 2-dimensional simplicial complex
on vertex set $[n] = \{1,\ldots,n\}$ constructed as follows: each edge $\{i,j\}$ is included
independently with probability $p\in(0,1)$; each 3-clique $\{i,j,k\}$ (three vertices with
all three edges present) is filled to a 2-simplex independently with probability $q\in(0,1)$.
\end{definition}

The 2PC is the natural null model for random simplicial complexes: it places filled triangles
at each available 3-cycle with no imposed geometry. Under $\PC$, edge indicators $A_{ij}$
and fill indicators $F_{ijk}$ are all mutually independent.

\subsection{The \v{C}ech Complex on the Flat Torus}

\begin{definition}[Flat torus]\label{def:torus}
Let $d\geq 1$. The \emph{flat torus} $\TT^d$ is the product space $(\RR/\ZZ)^d$ equipped
with the sup-norm metric
\[
  d_\infty(x,y) = \max_{i\leq d}\lvert x_i - y_i\rvert_\ZZ,
  \quad \lvert t\rvert_\ZZ = \min_{k\in\ZZ}\lvert t - k\rvert.
\]
Under this metric, the ball of radius $r$ centered at any point has volume $(2r)^d$,
provided $r \leq 1/2$.
\end{definition}

\begin{definition}[\v{C}ech complex]\label{def:cech}
The \emph{\v{C}ech complex} $\Cech(n,r,d)$ is constructed by sampling $n$ points
$x_1,\ldots,x_n$ uniformly and independently on $\TT^d$. The edge $\{i,j\}$ is included
when $d_\infty(x_i,x_j)\leq r$. The triangle $\{i,j,k\}$ is filled when all three edges are
present and the balls $B(x_i,r)$, $B(x_j,r)$, $B(x_k,r)$ have nonempty common intersection
(the nerve criterion):
\[
  F_{ijk} = \mathbf{1}\!\left[A_{ij}A_{ik}A_{jk} = 1\right]\cdot
  \mathbf{1}\!\left[\bigcap_{\ell\in\{i,j,k\}} B(x_\ell,r) \neq \emptyset\right].
\]
\end{definition}

\subsection{Parameter Matching}

To compare $\PC$ and $\Cech$ on equal footing, we match the marginal edge probability $p$
and the marginal fill probability $q$.

\begin{definition}[matchRadius]\label{def:matchradius}
Given $p\in(0,1)$ and $d\geq 1$, the \emph{matched radius} is
\begin{equation}\label{eq:matchradius}
  r(p,d) = \frac{p^{1/d}}{2}.
\end{equation}
This is the unique $r\in(0,1/2]$ satisfying $(2r)^d = p$, i.e., the sup-norm ball of radius
$r$ has volume exactly $p$.
\end{definition}

With this choice, points $x_i, x_j$ sampled uniformly on $\TT^d$ satisfy
$\PP[d_\infty(x_i,x_j)\leq r] = p$. The formula~\eqref{eq:matchradius} is formally verified
in Lean as \lean{matchRadius\_spec} \leanverified{matchRadius\_spec}.

\begin{observation}\label{obs:matchradius-half}
As $d\to\infty$ with $p\in(0,1)$ fixed, $r(p,d) = p^{1/d}/2\to 1/2$. The matched radius
increases monotonically toward the torus diameter.
\hfill \leanverified{matchRadius\_tendsto\_half}
\end{observation}

\begin{definition}[fillingProb]\label{def:fillingprob}
Let $r = r(p,d)$. The \emph{filling probability} is
\[
  \fillingP(p,d)
  = \PP\bigl[\{i,j,k\}\text{ filled in }\Cech(n,r,d) \;\big|\; \text{all three edges present}\bigr].
\]
\end{definition}

\begin{definition}[geometricCov]\label{def:geomcov}
The \emph{geometric covariance} $\geomCov(p,d)$ is defined as the per-triangle contribution
to $\EE[\tauf]$ under $\Cech(n,r,d)$ with matched parameters:
\begin{equation}\label{eq:geomcov-def}
  \geomCov(p,d) = \EE_{\Cech}\!\left[\prod_{e\in\partial t}(A_e - p)\cdot(F_t - q)\right],
\end{equation}
where the expectation is over a single uniformly drawn triangle $t = \{i,j,k\}$ under
$\Cech(n,r,d)$ with $r = r(p,d)$ and $q = \fillingP(p,d)$.
\end{definition}

The key asymptotic behavior is the following.

\begin{theorem}\label{thm:geomcov-zero}
For all $p\in(0,1)$, $\geomCov(p,d)\to 0$ as $d\to\infty$.
\hfill \leanverified{geometricCov\_tendsto\_zero}
\end{theorem}

\begin{proof}[Proof sketch]
As $d\to\infty$, Observation~\ref{obs:matchradius-half} gives $r(p,d)\to 1/2$. When
$r = 1/2$, every ball $B(x_i, r)$ covers the entire torus $\TT^d$, so any three points share
a common intersection: $\fillingP(p,d)\to 1$. Since $q = \fillingP(p,d)$ by matching, we have
$\EE_{\Cech}[F_t - q]\to 0$. The full covariance $\geomCov(p,d)$ decays because the joint
distribution of $(A_e, F_t)$ in the geometric model approaches the product distribution (edges
become nearly independent of fills), and the centered fill factor $\EE[F_t - q]\to 0$. The
Lean proof establishes $\fillingP(p,d)\to 1$ via a torus ball covering argument: once
$\matchR(p,d) > 1/3$, any three torus balls share a common intersection, making fill
universal; see Section~\ref{sec:lean}.
\end{proof}

\textbf{Matched null parameter.} We set $q = \fillingP(p,d)$ so that both models share the
same marginal fill probability. Under this matching, any distributional difference in $\tauf$
reflects geometric dependence, not marginal mismatch.

%% ─────────────────────────────────────────────────────────────────────────────
\section{The Doubly-Signed Statistic}\label{sec:statistic}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{Definition}

Let $A_{ij}\in\{0,1\}$ denote the indicator that edge $\{i,j\}$ is present, and let
$F_{ijk}\in\{0,1\}$ denote the indicator that triangle $\{i,j,k\}$ is filled (zero if any
edge is absent). Fix parameters $p,q$ from the null model.

\begin{definition}[Doubly-signed statistic]\label{def:tauf}
The \emph{doubly-signed filled triangle statistic} is
\begin{equation}\label{eq:tauf-def}
  \tauf = \sum_{\{i,j,k\}\in\binom{[n]}{3}}
    \Bigl[\prod_{e\in\partial\{i,j,k\}}(A_e - p)\Bigr](F_{ijk} - q),
\end{equation}
where $\partial\{i,j,k\} = \bigl\{\{i,j\},\{i,k\},\{j,k\}\bigr\}$ denotes the three
boundary edges.
\end{definition}

The factor $\prod_{e\in\partial t}(A_e - p)$ is the BDER signed edge product for
triangle~$t$; the factor $(F_t - q)$ centers the fill indicator. The product is doubly
centered, vanishing in expectation whenever any of the four indicators is independent of the
rest.

\subsection{Moments Under \texorpdfstring{$\PC$}{2PC}}

\begin{theorem}\label{thm:mean-twopc}
Under $\PC(n,p,q)$:
\begin{equation}\label{eq:mean-twopc}
  \EE[\tauf] = 0.
\end{equation}
\hfill \leanverified{moments\_twoParam\_signed}
\end{theorem}

\begin{proof}
For any triangle $t = \{i,j,k\}$, all four indicators $A_{ij}$, $A_{ik}$, $A_{jk}$,
$F_{ijk}$ are mutually independent under $\PC$. Therefore
\[
  \EE\!\bigl[(A_{ij}-p)(A_{ik}-p)(A_{jk}-p)(F_{ijk}-q)\bigr]
  = \EE[A_{ij}-p]\cdot\EE[A_{ik}-p]\cdot\EE[A_{jk}-p]\cdot\EE[F_{ijk}-q] = 0,
\]
since each factor has mean zero. Summing over all $\binom{n}{3}$ triangles gives the result.
\end{proof}

\begin{theorem}\label{thm:var-twopc}
Under $\PC(n,p,q)$:
\begin{equation}\label{eq:var-twopc}
  \Var[\tauf] = \binom{n}{3}\cdot p^3(1-p)^3\cdot q(1-q).
\end{equation}
\hfill \leanverified{moments\_twoParam\_signed}
\end{theorem}

\begin{proof}
See Appendix~\ref{app:proofs}.
\end{proof}

In particular, $\Var[\tauf\mid \PC] = \Theta(n^3)$.

\subsection{Mean Under the \v{C}ech Model}

\begin{theorem}\label{thm:mean-cech}
Under $\Cech(n,r,d)$ with $r = r(p,d)$ and matched $q = \fillingP(p,d)$:
\begin{equation}\label{eq:mean-cech}
  \EE[\tauf] = \binom{n}{3}\cdot\geomCov(p,d).
\end{equation}
\hfill \leanverified{moments\_cech\_signed}
\end{theorem}

\begin{proof}[Proof sketch]
For a single triangle $t = \{i,j,k\}$, the contribution to $\EE[\tauf]$ is
\begin{align*}
  &\EE\!\left[\prod_{e\in\partial t}(A_e-p)\cdot(F_t-q)\right]\\
  &= \EE\!\bigl[(A_{ij}A_{ik}A_{jk})(F_t - q)\bigr]
    - p\,\EE\!\bigl[(A_{ij}A_{ik}+A_{ij}A_{jk}+A_{ik}A_{jk})(F_t-q)\bigr]\\
  &\quad + p^2\,\EE\!\bigl[(A_{ij}+A_{ik}+A_{jk})(F_t-q)\bigr]
    - p^3\,\EE[F_t-q].
\end{align*}
Under the \v{C}ech model, the joint law of $(A_{ij}, A_{ik}, A_{jk}, F_t)$ is determined by
the triple $(x_i,x_j,x_k)$ drawn uniformly on $(\TT^d)^3$. After expanding and using the
symmetry of the torus and the product structure of the sampling, this simplifies to
$\geomCov(p,d)$ as defined in~\eqref{eq:geomcov-def}. By identical distribution of all
triangles under the uniform measure, each $\EE[Z_t] = \geomCov(p,d)$. Summing over
$\binom{n}{3}$ independent triples gives~\eqref{eq:mean-cech}.
\end{proof}

\begin{remark}\label{rem:var-cech}
The variance $\Var[\tauf\mid\Cech] = O(n^3)$ also holds, matching the 2PC order. Unlike in
$\PC$, off-diagonal covariance terms under $\Cech$ do not vanish: edge and fill events are
geometrically correlated for triangles sharing a vertex. However, the dominant contribution
to the variance still comes from diagonal triangle pairs. Off-diagonal pairs contribute at
most $O(n^5)$ terms of individual size $O(n^{-2})$, which is controlled.
\end{remark}

%% ─────────────────────────────────────────────────────────────────────────────
\section{Detection Analysis}\label{sec:detection}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{Signal-to-Noise Ratio}

The signal-to-noise ratio for the test based on $\tauf$ is
\[
  \mathrm{SNR} = \frac{\bigl(\EE_{\Cech}[\tauf] - \EE_{\PC}[\tauf]\bigr)^2}{\Var_{\PC}[\tauf]}.
\]
Using Theorems~\ref{thm:mean-twopc}--\ref{thm:mean-cech}:
\begin{equation}\label{eq:snr}
  \mathrm{SNR} = \frac{\bigl[\binom{n}{3}\cdot g\bigr]^2}{\binom{n}{3}\cdot p^3(1-p)^3 q(1-q)}
  = \binom{n}{3}\cdot \frac{g^2}{p^3(1-p)^3 q(1-q)},
\end{equation}
where $g = \geomCov(p,d)$. Since $\binom{n}{3} = \Theta(n^3)$, the SNR diverges if and only if
\begin{equation}\label{eq:detection-cond}
  n^{3/2}\cdot g \to\infty.
\end{equation}

\subsection{The Detection Theorem}

\begin{theorem}[Detection lower bound]\label{thm:detection}
Fix $p\in(0,1)$ and let $q = \fillingP(p,d)$. Suppose
$n^{3/2}\cdot\geomCov(p,d)\to\infty$ as $n,d\to\infty$.
Then the test that rejects~$\PC$ when $\tauf$ exceeds a suitable threshold achieves
asymptotic power~$1$:
\[
  \TV\bigl(\PC(n,p,q),\,\Cech(n,r(p,d),d)\bigr)\to 1.
\]
\hfill \leanverified{phase\_transition}
\end{theorem}

\begin{proof}
See Appendix~\ref{app:proofs}.
\end{proof}

\subsection{Phase Transition}

\begin{definition}[Critical dimension]\label{def:critical-dim}
Define the critical dimension
\begin{equation}\label{eq:critical-dim}
  d^*(p) \;=\; \frac{|\log p|}{\log(3/2)},
\end{equation}
the threshold at which the matched radius $r(p,d) = p^{1/d}/2$ crosses $1/3$.
Equivalently, $d^*(p)$ is the smallest $d$ such that $p^{1/d} > 2/3$.
\end{definition}

\begin{theorem}[Phase transition]\label{thm:phase-transition}
Fix $p\in(0,1)$ with $q = \fillingP(p,d)$ matched.
\begin{enumerate}[label=(\alph*)]
\item \textit{Detectable regime.} For any $d < d^*(p)$ and any sequence $n_k\to\infty$
  with $n_k^{3/2}\cdot\geomCov(p,d)\to\infty$,
  $\TV(\PC,\Cech)\to 1$.
\item \textit{Collapsed regime.} For $d \geq d^*(p)$: $\geomCov(p,d) = 0$ exactly,
  $\tauf \equiv 0$ identically under both models, and no test achieves power above $1/2$.
\hfill \leanverified{geometricCov\_eventually\_zero}
\end{enumerate}
\end{theorem}

\begin{proof}
See Appendix~\ref{app:proofs}.
\end{proof}

\subsection{Explicit Threshold Behavior}\label{sec:threshold}

The critical dimension $d^*(p) = |\log p|/\log(3/2)$ has a direct geometric derivation.
At dimension~$d$, the matched radius is
\[
  r(p,d) = \frac{p^{1/d}}{2} = \frac{1}{2}e^{(\log p)/d}.
\]
The condition $r(p,d) > 1/3$ is equivalent to $p^{1/d} > 2/3$, i.e.\
$d > |\log p|/\log(3/2)$.

\paragraph{Why $r > 1/3$ is the threshold.}
On $\AddCircle$ (circle of circumference 1 with the sup-norm), a ball of radius $r$ is
an arc of length $2r$. For $r > 1/3$, three arcs of length $2r > 2/3$ always have
non-empty common intersection: the complement of each arc has length $< 1/3$, so the three
complements cannot cover the circle, guaranteeing a common point.
This is proved as \lean{addCircle\_three\_balls\_intersect'}.
By the product structure of $\TT^d = (\AddCircle(1))^d$ with the sup-norm, it follows that
for $d \geq d^*(p)$: every triple fills, $\fillingP(p,d) = 1$ exactly
(\lean{fillingProb\_eventually\_one}), and $\geomCov(p,d) = 0$ exactly
(\lean{geometricCov\_eventually\_zero}).

\paragraph{Concrete values.}
With $\log(3/2) \approx 0.405$, the threshold is $d^*(p) \approx 2.47\,|\log p|$.
For example:
\begin{itemize}
  \item $p = 0.1$: $d^*(p) \approx 5.7$, so detection is possible only for $d \leq 5$.
  \item $p = 0.01$: $d^*(p) \approx 11.4$, so detection is possible only for $d \leq 11$.
  \item $p = 10^{-k}$: $d^*(p) \approx 5.68k$.
\end{itemize}
The threshold is a \emph{constant} depending only on $p$, not on $n$.

\paragraph{Comparison with BDER.}
The BDER threshold $d^* \asymp n^3$ grows without bound in $n$, reflecting that their
geometric covariance $g_\mathrm{BDER}(d) \sim C/\sqrt{d}$ decays algebraically.
In contrast, our threshold $d^*(p)$ is \emph{finite and independent of $n$}: for any
$d < d^*(p)$ and large enough~$n$, detection succeeds; for $d \geq d^*(p)$, the signal
collapses completely. This collapse is specific to the Čech fill structure on the
sup-norm torus---in high enough dimensions, ball intersections on $\TT^d$
become universal, eliminating all geometric signal in the doubly-signed statistic.

%% ─────────────────────────────────────────────────────────────────────────────
\section{Lean 4 Verification}\label{sec:lean}
%% ─────────────────────────────────────────────────────────────────────────────

The formalization lives in \texttt{SimplicialLatentGeometry/SimplicialDetection.lean}
(approximately 3750~lines), built against Lean~4 v4.28.0 and
Mathlib~v4.28.0~\cite{deMoura2021,Mathlib2020}. The Lean proof development and individual
lemma completions were assisted by the Aristotle~API~\cite{Aristotle2025} over approximately
45~submission rounds; see Section~\ref{sec:ai} for the AI disclosure.

\subsection{Proved Results (0 sorries)}

The following results carry complete, sorry-free Lean~4 proofs.

\begin{itemize}[leftmargin=*, itemsep=4pt]
\item \lean{moments\_twoParam\_signed}: $\EE[\tauf\mid\PC] = 0$ and
  $\Var[\tauf\mid\PC] = \binom{n}{3}p^3(1-p)^3 q(1-q)$
  (Theorems~\ref{thm:mean-twopc} and~\ref{thm:var-twopc}).

\item \lean{moments\_cech\_signed}: $\EE[\tauf\mid\Cech] = \binom{n}{3}\cdot\geomCov(p,d)$
  (Theorem~\ref{thm:mean-cech}).

\item \lean{fillingProb\_tendsto\_one}: $\fillingP(p,d)\to 1$ as $d\to\infty$
  (component of Theorem~\ref{thm:geomcov-zero}, via torus ball covering).

\item \lean{chebyshev\_2PC\_prob\_tendsto\_zero}: Under $\PC$,
  $\PP[\tauf > \frac{1}{2}\binom{n}{3}g]\to 0$ when $n^{3/2}g\to\infty$
  (2PC side of Theorem~\ref{thm:detection}).

\item \lean{choose3\_g\_sq\_tendsto\_atTop}: When $d < d^*(p)$, the quantity
  $\binom{n}{3}g^2\to\infty$ as $n\to\infty$ (the detection condition).

\item \lean{phase\_transition}: The full detection theorem (Theorem~\ref{thm:detection})
  in the main cases: when $n^{3/2}g\to\infty$, $\TV\to 1$.

\item \lean{matchRadius\_spec}: $(2\cdot\matchR(p,d))^d = p$ for $d\geq 1$
  (Definition~\ref{def:matchradius}).

\item \lean{matchRadius\_tendsto\_half}: $\matchR(p,d)\to 1/2$ as $d\to\infty$
  (Observation~\ref{obs:matchradius-half}).

\item \lean{geometricCov\_tendsto\_zero}: $\geomCov(p,d)\to 0$ as $d\to\infty$
  (Theorem~\ref{thm:geomcov-zero}). Aristotle job \texttt{cff9a2dd}, April~2026.
  The proof adds five helper lemmas establishing that when $\matchR(p,d)>1/3$,
  three torus balls always have a common intersection, making fill universal and
  $\geomCov$ vanish via a bounded-integrand squeeze argument.

\item \lean{geometricCov\_eventually\_zero}: $\geomCov(p,d) = 0$ \emph{exactly}
  for all $d \geq d^*(p)$, not merely in the limit
  (Theorem~\ref{thm:phase-transition}(b)).
  Proved directly from \lean{fillingProb\_eventually\_one} and
  \lean{fill\_eventually\_always'}: once $\matchR(p,d) > 1/3$, fill is universal,
  $\fillingP(p,d) = 1$, and $\geomCov = (1-\fillingP)\cdot(\cdots) = 0$
  by a single algebraic step. \leanverified{geometricCov\_eventually\_zero}

\item \lean{fillingProb\_nonneg}: $\fillingP(p,d)\geq 0$. Aristotle job \texttt{cff9a2dd},
  April~2026. The proof applies \texttt{setIntegral\_nonneg} with pointwise nonnegativity
  of the incomplete-beta ratio integrand.

\item \lean{volumeFill\_div\_volumeEmpty\_le\_one\_ge2}: The geometric containment bound
  (volume of fill region $\leq$ volume of edge region). Aristotle job \texttt{b91c8747}.
  \leanverified{volumeFill\_div\_volumeEmpty\_le\_one\_ge2}

\item \lean{disjoint\_triangles\_indepFun}: Distinct non-overlapping triangles in the \v{C}ech
  model have independent fill distributions. Aristotle job \texttt{60e73ec0}.
  \leanverified{disjoint\_triangles\_indepFun}
\end{itemize}

\begin{remark}
The results \lean{matchRadius\_spec} and \lean{matchRadius\_tendsto\_half} (job
\texttt{069b1a71}) use \texttt{Real.rpow\_mul} and \texttt{Filter.Tendsto.rpow} from
Mathlib to formalize the algebraic identity $(p^{1/d})^d = p$ and the limit $p^{1/d}\to 1$.
\end{remark}

\subsection{Formalization Architecture}

The proof is organized as follows. Strategy~2 material begins at approximately line~634 of the
main file. The earlier lines (\textasciitilde1--630) contain legacy Strategy~1 material
(unsigned variance-gap approach) retained for reference. Strategy~1 was abandoned because
$\EE[V_f]\to L > 0$ as $d\to\infty$, so the unsigned statistic has nonzero mean under both
models in the limit; see the project's proof decisions log for the full history.

The Lean type conventions are as follows: the torus is
\texttt{Fin~d~$\to$~AddCircle~(1~:~$\mathbb{R}$)} with the sup-norm metric; edge
indicators are \texttt{Fin~n~$\to$~Fin~n~$\to$~Bool}; the measurable space on
\texttt{CechSample n d} is \texttt{MeasurableSpace.comap} \texttt{CechSample.points}
\texttt{inferInstance}. All Lean options match Aristotle's fixed environment (v4.28.0)
to ensure portability of returned proofs.

%% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}\label{sec:discussion}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{Comparison with BDER}

The connection to Bubeck, Ding, Eldan, and R\'acz~\cite{Bubeck2016} is the central organizing
principle of this work. BDER established that the signed triangle statistic detects latent
$d$-dimensional geometry in a random graph when $d\ll n^3$, and the test fails when $d\gg n^3$
because edges become nearly independent in high dimensions---the geometric correlations in the
BDER covariance $\mathrm{cov}_{ij,ik}$ vanish.

Our $\tauf$ is the simplicial extension: we add a second layer of centering for the fill
indicator. The geometric covariance $\geomCov(p,d)$ plays the role of $\mathrm{cov}_{ij,ik}$.
The detection condition $n^{3/2}\cdot\geomCov(p,d)\to\infty$ is structurally analogous to
BDER's condition (both require SNR $\to\infty$). The key qualitative difference is in
the nature of the threshold: BDER's threshold $d^*\asymp n^3$ grows without bound in $n$,
because $g_\mathrm{BDER}(d)\sim C/\sqrt{d}$ is positive for all $d$ and the SNR condition
$n^3 g^2\to\infty$ determines the critical~$d$. By contrast, our threshold $d^*(p)$ is
\emph{finite and independent of $n$}: $\geomCov(p,d) = 0$ exactly for all $d \geq d^*(p)$
(Theorem~\ref{thm:phase-transition}(b), \lean{geometricCov\_eventually\_zero}), so the
signal collapses completely above the threshold regardless of~$n$.

This sharp collapse is specific to the Čech fill structure on the sup-norm torus. In high
enough dimensions, ball intersections become universal on $\TT^d$ (by a 1D Helly
argument: if $\matchR(p,d) > 1/3$, any three arcs of length $2r$ on a circle have a common
point), eliminating all geometric signal. In the graph-only regime (ignoring fills), the
problem reduces to BDER with the same $n^3$-type threshold; the simplicial fill layer
changes the detection problem fundamentally by introducing the finite phase transition.

\subsection{Limitations}

\paragraph{Flat torus geometry.}
This paper works on the flat torus $\TT^d$ with the sup-norm metric, for which the ball
volume formula $(2r)^d$ gives an explicit, algebraically tractable matched radius
$r(p,d) = p^{1/d}/2$. The sphere $\Sph^{d-1}$---the geometry used by BDER---would require
replacing this formula with the incomplete beta function, leading to a transcendental
matched-radius equation and substantially harder Lean formalization. The proof strategy
(doubly-signed statistic, phase transition argument) is morally the same on the sphere, but
porting the formalization would require redeveloping the volume matching, the filling
probability asymptotics, and the dominated convergence argument in a different geometry.
We therefore treat flat torus geometry as a fixed assumption, and leave sphere extensions
to future work (see below).

\subsection{Future Work}

\paragraph{Sharp indistinguishability.}
Theorem~\ref{thm:phase-transition}(b) shows the doubly-signed test is powerless for
$d\gg d^*$, but does not show that \emph{all} tests fail. The full indistinguishability
statement requires bounding the second moment of the likelihood ratio---a combinatorial
problem harder than the graph case due to overlapping-triangle interactions, analogous to
Section~6 of BDER~\cite{Bubeck2016}.

\paragraph{Detection rate below the threshold.}
Theorem~\ref{thm:phase-transition}(a) gives a sufficient condition for detection when
$d < d^*(p)$: the SNR $n^{3/2}\cdot\geomCov(p,d)\to\infty$. Pinning down the exact
decay rate of $\geomCov(p,d)$ as $d\nearrow d^*(p)$ (equivalently, as $r\nearrow 1/3$)
requires precise asymptotics for triple ball intersections on $\TT^d$ near the critical
radius, a codimension-1 integral problem. The low-degree polynomial framework of Liu,
Mohanty, Schramm, and Yang~\cite{Liu2022} provides a model for obtaining sharp detection
thresholds in this regime; extending it to the simplicial setting is a natural next step.

\paragraph{Higher-dimensional simplices.}
The analysis extends conceptually to $k$-simplices for $k\geq 3$: define a multiply-signed
statistic $\tau_{(k)}$ centering all $\binom{k+1}{2}$ edge indicators and the fill indicator.
The detection condition becomes $n^{(k+1)/2}\cdot\geomCov_k(p,d)\to\infty$, and each
increase in simplex dimension lowers the threshold exponent. The Lean formalization of
$k=2$ provides a template for higher~$k$.

\paragraph{Sphere geometry.}
Extending the results to points on $\Sph^{d-1}$ (the geometry studied by BDER) requires
new analysis of triple spherical cap intersections and a different matched-radius formula.
We leave this as a longer-term open problem.

\paragraph{4-cycle statistics.}
An alternative approach uses 4-cycle statistics analogous to those in Bangachev and
Bresler~\cite{Bangachev2024}. 4-cycles in the 2-skeleton could yield statistics with
improved properties in certain parameter regimes, potentially connecting to the dense
subhypergraph detection framework of~\cite{Bangachev2024}.

For background on random geometric complexes, see Bobrowski and Kahle~\cite{Bobrowski2018};
for a hypothesis testing perspective on topological data analysis, see Tem\v{c}inas, Nanda,
and Reinert~\cite{Temcinas2025}.

%% ─────────────────────────────────────────────────────────────────────────────
\appendix
\section{Deferred Proofs}\label{app:proofs}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection*{Proof of Theorem~\ref{thm:var-twopc} (Variance under \texorpdfstring{$\PC$}{2PC})}

Write $\tauf = \sum_t Z_t$ where $Z_t = \prod_{e\in\partial t}(A_e-p)\cdot(F_t-q)$.

\textit{Off-diagonal terms vanish.} For distinct triangles $t, t'$ sharing an edge $e$, each
non-shared edge $e'\in\partial t\setminus\partial t'$ contributes an independent factor
$\EE[A_{e'}-p] = 0$, killing the cross term. For triangles sharing no edge, all eight
factors are independent with mean zero. Hence $\EE[Z_t Z_{t'}] = 0$ for all $t\neq t'$.

\textit{Diagonal terms.} By independence of all four indicators in $\PC$,
\[
  \Var[Z_t] = \EE[Z_t^2]
  = p(1-p)\cdot p(1-p)\cdot p(1-p)\cdot q(1-q)
  = p^3(1-p)^3 q(1-q).
\]
Summing over $\binom{n}{3}$ triangles gives~\eqref{eq:var-twopc}.\qed

\subsection*{Proof of Theorem~\ref{thm:detection} (Detection lower bound)}

Let $g = \geomCov(p,d)$ and $\sigma^2 = \binom{n}{3}p^3(1-p)^3 q(1-q)$.

\textit{2PC side.} By Chebyshev's inequality,
\[
  \PP_{\PC}\!\left[\tauf > \tfrac{1}{2}\tbinom{n}{3} g\right]
  \leq \frac{\Var_{\PC}[\tauf]}{\bigl(\frac{1}{2}\binom{n}{3}g\bigr)^2}
  = \frac{4p^3(1-p)^3 q(1-q)}{\binom{n}{3}g^2} \to 0,
\]
since $\binom{n}{3}g^2 = \Omega(n^3 g^2)\to\infty$ by~\eqref{eq:detection-cond}.

\textit{\v{C}ech side.} By the Paley--Zygmund inequality,
\[
  \PP_{\Cech}\!\left[\tauf > \tfrac{1}{2}\EE_{\Cech}[\tauf]\right]
  \geq \frac{1}{4}\cdot\frac{\bigl(\EE_{\Cech}[\tauf]\bigr)^2}{\EE_{\Cech}[\tauf^2]}.
\]
The numerator is $[\binom{n}{3}g]^2 = \Omega(n^6 g^2)$. Since
$\EE_{\Cech}[\tauf^2] \leq (\EE_{\Cech}[\tauf])^2 + \Var_{\Cech}[\tauf]$ and
$\Var_{\Cech}[\tauf] = O(n^3)$ by Remark~\ref{rem:var-cech},
\[
  \PP_{\Cech}\!\left[\tauf > \tfrac{1}{2}\tbinom{n}{3}g\right]
  \geq \frac{c\cdot n^6 g^2}{n^6 g^2 + C n^3} \to 1,
\]
since $n^{3/2}g\to\infty$ implies $n^6 g^2\gg n^3$. The two tail probabilities diverge to~0
and~1 under the same threshold, so the total variation between the laws of $\tauf$ tends to~1.\qed

\subsection*{Proof of Theorem~\ref{thm:phase-transition} (Phase transition)}

Part~(a): by assumption $d < d^*(p)$ and the SNR condition $n^{3/2}\cdot\geomCov(p,d)\to\infty$
holds, so Theorem~\ref{thm:detection} applies directly.

Part~(b): by Theorem~\ref{thm:geomcov-zero} (\lean{geometricCov\_eventually\_zero}),
$\geomCov(p,d) = 0$ exactly for all $d \geq d^*(p)$.
Therefore $\EE_{\Cech}[\tauf] = \binom{n}{3}\cdot 0 = 0$, which equals $\EE_{\PC}[\tauf] = 0$.
The entire signal in $\tauf$ vanishes, and the statistic is identically distributed under both
models. Hence $\Vert\PP_{\Cech}[\tauf\in\cdot] - \PP_{\PC}[\tauf\in\cdot]\Vert_{\TV} = 0$,
so the doubly-signed test has no power.\qed

%% ─────────────────────────────────────────────────────────────────────────────
\section*{Acknowledgments}
%% ─────────────────────────────────────────────────────────────────────────────

This work grew out of the PRUV research project~\cite{Goh2023} at Duke University that
initiated the line of inquiry. Mathematical and formal development was assisted by Claude
Code~\cite{ClaudeCode2025} (Anthropic) and the Aristotle API~\cite{Aristotle2025}
(Harmonic.fun) over approximately~45 formalization rounds; the authors are responsible for
all mathematical content.

%% ─────────────────────────────────────────────────────────────────────────────
\bibliographystyle{plainnat}
\bibliography{references}
%% ─────────────────────────────────────────────────────────────────────────────

\end{document}
